---
title: "Homework 11"
author: "Annanina Koster"
date: "15 december 2015"
output: html_document
---

The homework [17 XII]
For the eunomia study perform following analyses:

Try a post-hoc test for effect of ICD10 to find out which diseases are more likely to produce psychotic behaviours?
Is it possible you cluster ICD10 groups depending on average level of psychotic behaviours?
Is it possible to do something similar if ICD10 is coded as random effects?

---

The purpose of this homework assignment is to investigate the relation between diseases and the level of psychotic behaviour in the EUNOMIA dataset. First, I open the data and take a look at what I've got. I decide on using the BPRS.T2 variable (BPRS - Average score for Brief Psychiatric Rating Scale, measured in times: T1, T2 and T3) to represent the level of psychotic behaviour. I chose it at time T2, because I think that half a year after treatment the most interesting results can be obtained. ICD10 - International Statistical Classification of Diseases and Related Health Problems 10th Revision, was chosen to represent the diseases.

```{r, warning=FALSE, message=FALSE}

library(PBImisc)
summary(eunomia[,c(1:5,9:11,15)])

```

As can be seen, a lot of different diseases are being considered in this study. Hence, plots will not be very informative. Therefore, a Scheffe post-hoc test for ICD10 will give me more information when ICD10 is modeled as a fixed effect. 

```{r, warning=FALSE, message=FALSE}

model.fixed <- lm(BPRS.T2 ~ ICD10 + NUM.HOSP, data = eunomia) #GENDER and AGE are not significant when added and CENTRE13 should be coded as a random effect. NUMP.HOSP is significant, so should be added!
library(agricolae)
head(scheffe.test(aov(model.fixed), 'ICD10', console = FALSE)$groups)

```

According to Scheffe test, for all diseases the average level of psychotic behaviours are equal (they are all in group a). Thus, it is not possible to cluster ICD10 groups depending on average level of psychotic behaviours. It would be clustered as 1 cluster.

IF ICD10 is coded as a random effect, only the variance of these diseases compared to the level of psychotic behaviour is known. Hence, ICD10 cannot be clustered into groups with similar averages of BPRS.T2.

```{r, warning=FALSE, message=FALSE}

model.random1 <- lmer(BPRS.T2 ~ NUM.HOSP + (1|ICD10) + (NUM.HOSP|CENTRE13), data = eunomia, REML = FALSE) #this model was found to be best in the lab.
model.random2 <- lmer(BPRS.T2 ~ NUM.HOSP + (NUM.HOSP|CENTRE13), data = eunomia, REML = FALSE)
anova(model.random1, model.random2)

```

This Likelihood Ratio test shows that the random effect ICD10 is indeed signifcant and should be included in the model as the variable it is now, and not as 1 cluster!

