---
title: "Homework 11"
author: "Agnieszka Sitko"
output:
  html_document:
    toc: yes
---

```{r, warning = FALSE, message = FALSE, echo = FALSE}
library(lattice)
library(PBImisc)
library(lme4)
attach(sleepstudy)
      
```
##Plots

The relation between `Reaction` and `Days` for each subject is plotted below. 

```{r, warning = FALSE, message = FALSE, echo = FALSE}

xyplot(Reaction ~ Days | Subject, type = c("g","p","r"),
      xlab = "Days", ylab = "Reaction", pch = 19,
      panel = lattice.getOption("panel.xyplot"), col = "blue3")
```

For each plot linear trendline is added. Most subjects tend to increase their reaction time with the increase in the number of days with restricted amount of sleep. 

##Model

We will introduce a model which assumes that there is a linear and fixed relation between `Reaction` and `Days` in a population as a whole and there may exist random effect of `Subject`. The results of such a model are listed as follows.

```{r, warning = FALSE, message = FALSE, echo = FALSE}
model <- lmer(Reaction ~ Days + (Days|Subject))
summary(model)
```

From this summary we can say that there exists positive relation between the number of days with sleep restriction and reaction time of a subject in a population.

It may be shown that the model is correct (all variables are important, random variables are normally distributed).

SIGNIFICANCE OF THE FIXED EFFECT 
```{r, warning = FALSE, message = FALSE}
2 * pnorm(6.77, lower.tail = FALSE)
```

SIGNIFICANCE OF THE RANDOM EFFECT
```{r, warning = FALSE, message = FALSE}
model1 <- lm(Reaction ~ Days)
anova(model, model1)[[8]][2]
```

MODEL DIAGNOSTICS
```{r, warning = FALSE, message = FALSE, echo = FALSE}
ranEf <- ranef(model, postVar = TRUE)
qqmath(ranEf)
qqnorm(fitted.values(model)-Reaction, main = "QQ plot for residuals"); qqline(fitted.values(model)-Reaction)
```

##Individual slopes

Main statistics of slopes calculated for the model may be find below.
```{r, warning = FALSE, message = FALSE}
slopes <- ranEf$Subject[,2] + fixef(model)[2]
summary(slopes)
```

As expected most slopes are positive (only one is negative). So, even if we consider the random effect of subjects, the positive relation between `Reaction` and `Days` will be kept.