---
title: "Homework 11"
author: "Ma³gorzata Pujszo"
date: "Saturday, January 02, 2016"
output: html_document
---

I considered data sleepstudy from lme4 package. Firstly, let's see what our data looks like.


```{r, message=FALSE, warning=FALSE}
library(lme4)
library(lattice)
library(ggplot2)
head(sleepstudy)
nrow(sleepstudy)
unique(sleepstudy$Subject)
length(unique(sleepstudy$Subject))
```

As we can see our data is in a long format - we have information of reaction of a given patient in a given day. As a result, one patient is present in a data frame a couple times.

Now, I plotted the relation between Reaction and Days for each Subject independently.

```{r}
panel.with.square <- function(...){
  nx = list(...)$x
  ny = list(...)$y
  cc = lm(ny~nx)$coef
  panel.xyplot(...)
  panel.curve(x*cc[2]+cc[1], from=min(nx), to=max(nx), col="red")
}

xyplot(Reaction ~ Days | Subject, sleepstudy, type = c("g","p","r"),
  xlab = "Days", ylab = "Reaction", pch=19,
  panel = panel.with.square)
```

From the plot, one can infer that slopes for each Subject are different. Building a model we should take it into account. Consequently, I modelled the relation between Reaction and Days as linear but I also added random component depending on Subject. 

```{r}
model <- lmer(Reaction ~ Days + (Days|Subject), data=sleepstudy)
summary(model)
```

In the model, I allowed occurence of dependence of random effects. Let's test whether this assumption of dependence is significant.

```{r}
model1 <- lmer(Reaction ~ Days + (Days-1|Subject) + (1|Subject), data=sleepstudy)
summary(model1)
anova(model, model1)
```

Since the p-value is high, we don't need to assume dependence of random effects. In further analysis I used model1.

The last thing I did was extracting slopes for different patients (there are in vector "slopes") and presenting summary statistics.

```{r}
slopes <- ranef(model1)$Subject[,1] + fixef(model1)[2]
summary(slopes)
```

