---
title: "Project 1, Phase 1"
author: "Agnieszka Sitko, Annanina Koster, Rafał Cylwa, Piotr Obarski"
date: "Modele liniowe i mieszane"
output: 
  html_document:
  toc: TRUE
---
# Preparing data to analysis

Firstly we launch load library agricolae which is needed to perform post-hoc test used later and we load data.
```{r}
library(agricolae)
library(MASS)
load("~/Modele liniowe i mieszane/projekt 1/clinical.cb.rda")
load("~/Modele liniowe i mieszane/projekt 1/expression.cb1.rda")
load("~/Modele liniowe i mieszane/projekt 1/expression.cb2.rda")
```
We merge data contained in files expression.cb1 and expression.cb2, and then standarize codes of patients in file clinical.cb in order to have the same type of names as in file expression. 
```{r}
expression <- rbind(expression.cb1, expression.cb2)
remove(expression.cb1, expression.cb2)
clinical.cb$sampleID<-sub("-", ".", clinical.cb$sampleID)
clinical.cb$sampleID<-sub("-", ".", clinical.cb$sampleID)
clinical.cb$sampleID<-sub("-", ".", clinical.cb$sampleID)
```
Then we change the names of rows in expression

```{r}
rownames(expression)<- expression$Sample
expression <- expression[, -1]

t_expression <- t(expression)
t_expression <- as.data.frame(t_expression)

colnames(clinical.cb)[1]<-"id"
t_expression$id<- rownames(t_expression)
data<- merge(clinical.cb, t_expression, by="id")
```
# Analyzing data

## Checking assumptions
The results of a one-way ANOVA can be considered reliable as long as the following assumptions are met:<br/>

Response variable residuals are normally distributed (or approximately normally distributed).<br/>
Samples are independent.<br/>
Variances of populations are equal.<br/>
Responses for a given group are independent and identically distributed normal random variables (not a simple random sample (SRS)).<br/>
ANOVA is a relatively robust procedure with respect to violations of the normality assumption.

```{r}
colnames(data)[18] <- "CancerType"

normality<-1
CancerType <- data$CancerType
for (i in 30:length(colnames(data))) {
  x <- data[,i]
  model = aov(x ~ CancerType)
  normality[i]<-(shapiro.test(rstandard(model))$p.value>=0.01)
}

which(normality==FALSE)
length(which(normality==FALSE))/(i-30)

```
Assumption about normality of residuals is not fulfilled, so we will transform our data.
# Data transformation
```{r}

for(i in 30:length(colnames(data))) {
  x <- data[,i]
  x<-x+1-min(x) # shift data to get only positive
  a<-boxcox(x~CancerType)
  if (a$x[which.max(a$y)]!= 0) {
    x<-(x^(a$x[which.max(a$y)])-1)/a$x[which.max(a$y)]
  } else {
    x<-log(x)
  } 
  data[,i]<-x
}
```


in Scheffe test we do not need equal numbers of observations in groups as well as in LSD test, but we have many groups, and Scheffe is more conservative, so we will get less number of genes, what I think is quite good, because ultimately such a quantititave analysis leads to quality analysis
```{r}
normality<-1
normality_in_groups<-1
variance_of_groups<-1
for (i in 30:length(colnames(data))) {
  x <- data[,i]
  variance_in_groups[i]<-(bartlett.test(x, CancerType)$p.value>=0.01)
  zmienna<-0
  for (j in 1:length(levels(CancerType))){
    zmienna[j]<-tapply(x, CancerType, shapiro.test)[[j]]$p.value
  }
  normality_in_groups[i]<-(min(zmienna)>=0.05)
  model = anova(lm(x ~ CancerType))
  if(model$"Pr(>F)"[1]<=0.05){
  nrofgroups[i]<-match(tail(scheffe.test(model, "CancerType", console = FALSE)$groups$M,1), letters)
  }
  normality[i]<-(shapiro.test(rstandard(model))$p.value>=0.01)
}
which(normality==FALSE)
length(which(normality==FALSE))/(i-30)

which(normality_in_groups==FALSE)
length(which(normality_in_groups==FALSE))/(i-30)
which(variance_in_groups==FALSE)
length(which(variance_in_groups==FALSE))/(i-30)
```


## Conclusions

```{r}
hist(nrofgroups[30:length(nrofgroups)])
colnames(data)[nrofgroups>7 & (is.na(nrofgroups) == FALSE)]
```