---
title: "Homework 5"
author: "Agnieszka Sitko"
output:
  html_document:
    toc: yes
  pdf_document:
    toc: yes
---
We will examine values of systolic blood pressure under the assumption that the mean systolic blood pressure for women diagnosed with diabetes type 1 or prediabetes is 120. 
Let's assume that in our samples for each 100 men there are around 106.5 women (using Polish feminization rate). 
Types of diabetes are assigned with the probabilities mentioned in the study: http://www.mz.gov.pl/__data/assets/pdf_file/0004/26851/Krzysztof-Strojek.pdf. 

```{r, warning = FALSE, message = FALSE}

library(agricolae)

set.seed(12)

pvals <- 0
power <- 0
M <- 200

for (N in 50: 1000){ 
      
      pvals <- replicate(M, {
            
            #generating dataset
            mean_systolic_blood_pressure <- rep(120, times = N)
            epsilon <- rnorm(N)
            gender <- sample(c("men", "women"), N, replace = TRUE, prob = c(0.484, 0.516))
            diabetes <- sample( c("pre", "1", "2"),  N, replace = TRUE, prob = c(0.52, 0.05, 0.43))
            
            #creating a model for the decribed scenario
            real_systolic_blood_pressure <- mean_systolic_blood_pressure + 
                  (gender == "men") * 1 + (gender == "men") * (diabetes == "2") * 0.5 + 
                  (gender == "women") * (diabetes == "2") * 1 + epsilon
            model <- lm(real_systolic_blood_pressure ~ gender / diabetes)
            
            summary(aov(model))[[1]][[5]][2] })
      
      power[N] <- sum(pvals < 0.05) / M
            
         
}


plot(power, main = "Power function", xlab = "population size", ylab = "power of a test for nested effects")
abline(h = 0.8, col = "blue")
abline(v = max(which(power < 0.8)) + 1 , col = "blue")
```


As it is seen in the plot, power as a function of a sample size is not stricly increasing. However, it fluctuates around the positive trend. So, for a minimal number of patients that allow to detect effect of the size (0,0,0.5,0,0,1) with probability larger than 80% I will take the number $N$ such that for any $n \geq  N$ the power of a test for nested effects is greater than 0.8.
```{r, message= FALSE, warning=FALSE}

max(which(power < 0.8)) + 1 


```
