---
title: "Untitled"
author: "Emilia Pompe"
date: "Wednesday, November 04, 2015"
output: html_document
---
### Emilia Pompe
### Homework 5

I assume that gender and diabetes type are independent random variables. The function dataSimulation creates a dataset of n observations and 3 variables: gender, diabetes type and blood pressure. The dataset is created in a way to follow the instructions given in the homework description. There are 3 argument of this function: n - number of observations, probs.gender - probabilities of "woman"/ "man" set by default to NULL and probs.diabetes - probabilities of simulating each diabetes type (also set by default to NULL).

```{r}
library(ggplot2)

dataSimulation <- function(n, probs.gender = NULL, probs.diabetes = NULL){
  gender <- sample(c("man", "woman"), size = n, replace = TRUE, prob = probs.gender)
  diabetes <- sample(c("pre", "1", "2"), size = n, replace = TRUE, prob = probs.diabetes)
  dataset <- data.frame(gender = factor(gender), diabetes = factor(diabetes))
  dataset$blood.pressure <- 120 + rnorm(n)
  dataset$blood.pressure <- dataset$blood.pressure + 1*(gender == "man") + 
    1*(gender == "woman" & diabetes == "2") + 0.5*(gender == "man" & diabetes == "2")
  return(dataset)
}
```

The function below, testPower depends on 5 parameters: n, probs.gender and probs.diabetes - as in the function dataSimulation, and alpha - significance level, N - how many times the experiment is repeated. The function creates a dataset with n observations, as it was explained before. Then, it calculates p-value of the ANOVA test for the nested effect. The procedure is repeated N times and empirical power of this test is calculated as the percentage of p-values lower than alpha.
```{r}
testPower <- function(N, alpha, n, probs.gender = NULL, probs.diabetes = NULL){
 p.values <- replicate(N, {
    dataset <- dataSimulation(n, probs.gender = probs.gender, probs.diabetes = probs.diabetes)
    model <- aov(blood.pressure ~ gender/diabetes, data = dataset)
    summary(model)[[1]][[5]][2]
    })
  sum(p.values < alpha)/N
}
```

I used the function to find the minimal sample size needed to obtain the test power greater than or equal to 0.8. I created 4 different scenarios. In first two of them all  3 diabetes types are equally probable in population. The difference is in the significance level (0.05 or 0.01).
```{r}
set.seed(7)

alpha = 0.05
N = 1000
powers<- sapply(20:220, function(i){
  testPower(N, alpha, i)
})
```

The minimal required sample size is in this case equal to:
```{r}
min(19+(which(powers>=0.8)))
qplot(x = 20:220, y = powers) + geom_line() + xlab("sample size") +
  ylab("test power") + geom_abline(intercept = 0.8, slope = 0, color = "red") +
  ggtitle("Test power: alpha = 0.05, diabetes types equally probable")
```

The same analysis for different significance level.
```{r}
alpha = 0.01
N = 1000
powers<- sapply(20:220, function(i){
  testPower(N, alpha, i)
})
```

The minimal required sample size is in this case equal to:
```{r}
min(19+(which(powers>=0.8)))
qplot(x = 20:220, y = powers) + geom_line() + xlab("sample size") +
  ylab("test power") + geom_abline(intercept = 0.8, slope = 0, color = "red") +
  ggtitle("Test power: alpha = 0.01, diabetes types equally probable")
```

Clearly the minimal required sample size is higher. This result seems quite logical: probability of rejecting the null hypothesis is smaller, hence the power of the test is lower for a given sample size. It is worth noting that it may be observed from the plots that the higher the sample size, the greater the test power.
I tried the same analysis but with different probabilities vector for the diabetes type.
```{r}
set.seed(7)

alpha = 0.05
N = 1000
powers<- sapply(20:220, function(i){
  testPower(N, alpha, i, probs.diabetes = c(0.5, 0.3, 0.2))
})
```

The minimal required sample size is in this case equal to:
```{r}
min(19+(which(powers>=0.8)))
qplot(x = 20:220, y = powers) + geom_line() + xlab("sample size") +
  ylab("test power") + geom_abline(intercept = 0.8, slope = 0, color = "red") +
  ggtitle("Test power: alpha = 0.05, diabetes types not equally probable")
```

Let us change again the significance level.
```{r}
alpha = 0.01
N = 1000
powers<- sapply(20:220, function(i){
  testPower(N, alpha, i, probs.diabetes = c(0.5, 0.3, 0.2))
})
```

The minimal required sample size is in this case equal to:
```{r}
min(19+(which(powers>=0.8)))
qplot(x = 20:220, y = powers) + geom_line() + xlab("sample size") +
  ylab("test power") + geom_abline(intercept = 0.8, slope = 0, color = "red") +
  ggtitle("Test power: alpha = 0.01, diabetes types not equally probable")
```

The conclusion from the results are similar to those obtained in case of equally probable diabetes types.