---
title: "Homework 7"
author: "Annanina Koster"
date: "17 november 2015"
output: html_document
---

Home work for 19 XI 2015
Read the chapter 2.6.3 from ,,Analiza danych z programem R, Modele liniowe mieszane'' or in English (http://www.stat.umn.edu/geyer/5931/mle/sel.pdf).

Consider three approaches to variable selection:

choose model with best BIC,
choose model with best AIC,
choose model with significant variables (i.e. marginal F test).
Generate three scenarios in such a way that in the first scenario the BIC will choose the right model in most cases, in the second scenario the AIC will be the best one and in the third scenario the selection with significant variables will find the right model most often.

In all scenarios the number of variables should be equal 10. The number of observations is up to you. The noise should be generated from N(0,1) distribution. For each variable you should decide if the variable effect is equal to 0 (if variable is not related with outcome) of is different than 0 (variable is important).

In your homework present results from 100 (or more) replications. For each criteria calculate the fraction of times that the criteria choose the right model.

- - -

For the first situation, I will choose quite a small number of observations (n), because the BIC model is then more likely to select the correct model (because the value will be small). 

```{r}

situation1 <- function(n, imp.par){

par <- 10
# coefficients
beta <- c(rep(1,imp.par),rep(0,par-imp.par))

# variables
X <- matrix(rnorm(n*par),n,par)
colnames(X) <- paste0("var", 1:par)

# outcome variable
y <- X %*% t(t(beta)) + rnorm(n)

dat <- data.frame(y,X)

# let's see the model
model1.0 <- lm(y ~ 1, data=dat)
model1.1 <- lm(y ~ var1, data=dat)
model1.2 <- lm(y ~ var1 + var2, data=dat)
model1.3 <- lm(y ~ var1 + var2 + var3, data=dat)
model1.4 <- lm(y ~ var1 + var2 + var3 + var4, data=dat)
model1.5 <- lm(y ~ var1 + var2 + var3 + var4 + var5, data=dat)
model1.6 <- lm(y ~ var1 + var2 + var3 + var4 + var5 + var6, data=dat)
model1.7 <- lm(y ~ var1 + var2 + var3 + var4 + var5 + var6 + var7, data=dat)
model1.8 <- lm(y ~ var1 + var2 + var3 + var4 + var5 + var6 + var7 + var8, data=dat)
model1.9 <- lm(y ~ var1 + var2 + var3 + var4 + var5 + var6 + var7 + var8 + var9, data=dat)
model1.10 <- lm(y ~ var1 + var2 + var3 + var4 + var5 + var6 + var7 + var8 + var9 + var10, data=dat)

BIC1.0 <- BIC(model1.0)
BIC1.1 <- BIC(model1.1)
BIC1.2 <- BIC(model1.2)
BIC1.3 <- BIC(model1.3)
BIC1.4 <- BIC(model1.4)
BIC1.5 <- BIC(model1.5)
BIC1.6 <- BIC(model1.6)
BIC1.7 <- BIC(model1.7)
BIC1.8 <- BIC(model1.8)
BIC1.9 <- BIC(model1.9)
BIC1.10 <- BIC(model1.10)

AIC1.0 <- AIC(model1.0)
AIC1.1 <- AIC(model1.1)
AIC1.2 <- AIC(model1.2)
AIC1.3 <- AIC(model1.3)
AIC1.4 <- AIC(model1.4)
AIC1.5 <- AIC(model1.5)
AIC1.6 <- AIC(model1.6)
AIC1.7 <- AIC(model1.7)
AIC1.8 <- AIC(model1.8)
AIC1.9 <- AIC(model1.9)
AIC1.10 <- AIC(model1.10)

CP1.0 <- sum(model1.0$residuals^2)/(summary(model1.10)$sigma^2) + 2 * model1.0$rank -
+ 100
CP1.1 <- sum(model1.1$residuals^2)/(summary(model1.10)$sigma^2) + 2 * model1.1$rank -
+ 100
CP1.2 <- sum(model1.2$residuals^2)/(summary(model1.10)$sigma^2) + 2 * model1.2$rank -
+ 100
CP1.3 <- sum(model1.3$residuals^2)/(summary(model1.10)$sigma^2) + 2 * model1.3$rank -
+ 100
CP1.4 <- sum(model1.4$residuals^2)/(summary(model1.10)$sigma^2) + 2 * model1.4$rank -
+ 100
CP1.5 <- sum(model1.5$residuals^2)/(summary(model1.10)$sigma^2) + 2 * model1.5$rank -
+ 100
CP1.6 <- sum(model1.6$residuals^2)/(summary(model1.10)$sigma^2) + 2 * model1.6$rank -
+ 100
CP1.7 <- sum(model1.7$residuals^2)/(summary(model1.10)$sigma^2) + 2 * model1.7$rank -
+ 100
CP1.8 <- sum(model1.8$residuals^2)/(summary(model1.10)$sigma^2) + 2 * model1.8$rank -
+ 100
CP1.9 <- sum(model1.9$residuals^2)/(summary(model1.10)$sigma^2) + 2 * model1.9$rank -
+ 100
CP1.10 <- sum(model1.10$residuals^2)/(summary(model1.10)$sigma^2) + 2 * model1.10$rank -
+ 100

values1 <- c(BIC1.0, BIC1.1, BIC1.2, BIC1.3, BIC1.4, BIC1.5, BIC1.6, BIC1.7, BIC1.8, BIC1.9, BIC1.10, AIC1.0, AIC1.1, AIC1.2, AIC1.3, AIC1.4, AIC1.5, AIC1.6, AIC1.7, AIC1.8, AIC1.9, AIC1.10, CP1.0, CP1.1, CP1.2, CP1.3, CP1.4, CP1.5, CP1.6, CP1.7, CP1.8, CP1.9, CP1.10)

return(values1)

}

set.seed(124)

imp.par <- 5

count1 <- replicate(100, situation1(100, imp.par))

count1.bic <- matrix(0,11,100)
count1.aic <- matrix(0,11,100)
count1.cp <- matrix(0,11,100)
for(i in 1:100){
  for(j in 1:11){
    count1.bic[j,i] <- count1[j,i]
  }
  for(j in 12:22){
    count1.aic[j-11,i] <- count1[j,i]
  }
  for(j in 23:33){
    count1.cp[j-22,i] <- count1[j,i]
  }
}

best.model.bic <- 0
best.model.aic <- 0
best.model.cp <- 0
for(i in 1:100){
  best.model.bic[i] <- which.min(count1.bic[,i]) #BIC
  best.model.aic[i] <- which.min(count1.aic[,i]) #AIC
  best.model.cp[i] <- which.min(count1.cp[,i]) #marginal F test
}

count.bic <- 0
count.aic <- 0
count.cp <- 0

for(i in 1:100){
  if(best.model.bic[i]==(imp.par+1)){
    count.bic <- count.bic + 1
  }
  if(best.model.aic[i]==(imp.par+1)){
    count.aic <- count.aic + 1
  }
  if(best.model.cp[i]==(imp.par+1)){
    count.cp <- count.cp + 1
  }
}

count.bic #97
count.aic #75
count.cp #78


```

AIC will be a better model than BIC if n is large.

```{r}

situation2 <- function(n, imp.par){

par <- 10
# coefficients
beta <- c(rep(1,imp.par),rep(0,par-imp.par))

# variables
X <- matrix(rnorm(n*par,0,0.001),n,par)
colnames(X) <- paste0("var", 1:par)

# outcome variable
y <- X %*% t(t(beta)) + rnorm(n)

dat <- data.frame(y,X)

# let's see the model
model1.0 <- lm(y ~ 1, data=dat)
model1.1 <- lm(y ~ var1, data=dat)
model1.2 <- lm(y ~ var1 + var2, data=dat)
model1.3 <- lm(y ~ var1 + var2 + var3, data=dat)
model1.4 <- lm(y ~ var1 + var2 + var3 + var4, data=dat)
model1.5 <- lm(y ~ var1 + var2 + var3 + var4 + var5, data=dat)
model1.6 <- lm(y ~ var1 + var2 + var3 + var4 + var5 + var6, data=dat)
model1.7 <- lm(y ~ var1 + var2 + var3 + var4 + var5 + var6 + var7, data=dat)
model1.8 <- lm(y ~ var1 + var2 + var3 + var4 + var5 + var6 + var7 + var8, data=dat)
model1.9 <- lm(y ~ var1 + var2 + var3 + var4 + var5 + var6 + var7 + var8 + var9, data=dat)
model1.10 <- lm(y ~ var1 + var2 + var3 + var4 + var5 + var6 + var7 + var8 + var9 + var10, data=dat)

BIC1.0 <- BIC(model1.0)
BIC1.1 <- BIC(model1.1)
BIC1.2 <- BIC(model1.2)
BIC1.3 <- BIC(model1.3)
BIC1.4 <- BIC(model1.4)
BIC1.5 <- BIC(model1.5)
BIC1.6 <- BIC(model1.6)
BIC1.7 <- BIC(model1.7)
BIC1.8 <- BIC(model1.8)
BIC1.9 <- BIC(model1.9)
BIC1.10 <- BIC(model1.10)

AIC1.0 <- AIC(model1.0)
AIC1.1 <- AIC(model1.1)
AIC1.2 <- AIC(model1.2)
AIC1.3 <- AIC(model1.3)
AIC1.4 <- AIC(model1.4)
AIC1.5 <- AIC(model1.5)
AIC1.6 <- AIC(model1.6)
AIC1.7 <- AIC(model1.7)
AIC1.8 <- AIC(model1.8)
AIC1.9 <- AIC(model1.9)
AIC1.10 <- AIC(model1.10)

CP1.0 <- sum(model1.0$residuals^2)/(summary(model1.10)$sigma^2) + 2 * model1.0$rank -
+ 100
CP1.1 <- sum(model1.1$residuals^2)/(summary(model1.10)$sigma^2) + 2 * model1.1$rank -
+ 100
CP1.2 <- sum(model1.2$residuals^2)/(summary(model1.10)$sigma^2) + 2 * model1.2$rank -
+ 100
CP1.3 <- sum(model1.3$residuals^2)/(summary(model1.10)$sigma^2) + 2 * model1.3$rank -
+ 100
CP1.4 <- sum(model1.4$residuals^2)/(summary(model1.10)$sigma^2) + 2 * model1.4$rank -
+ 100
CP1.5 <- sum(model1.5$residuals^2)/(summary(model1.10)$sigma^2) + 2 * model1.5$rank -
+ 100
CP1.6 <- sum(model1.6$residuals^2)/(summary(model1.10)$sigma^2) + 2 * model1.6$rank -
+ 100
CP1.7 <- sum(model1.7$residuals^2)/(summary(model1.10)$sigma^2) + 2 * model1.7$rank -
+ 100
CP1.8 <- sum(model1.8$residuals^2)/(summary(model1.10)$sigma^2) + 2 * model1.8$rank -
+ 100
CP1.9 <- sum(model1.9$residuals^2)/(summary(model1.10)$sigma^2) + 2 * model1.9$rank -
+ 100
CP1.10 <- sum(model1.10$residuals^2)/(summary(model1.10)$sigma^2) + 2 * model1.10$rank -
+ 100

values1 <- c(BIC1.0, BIC1.1, BIC1.2, BIC1.3, BIC1.4, BIC1.5, BIC1.6, BIC1.7, BIC1.8, BIC1.9, BIC1.10, AIC1.0, AIC1.1, AIC1.2, AIC1.3, AIC1.4, AIC1.5, AIC1.6, AIC1.7, AIC1.8, AIC1.9, AIC1.10, CP1.0, CP1.1, CP1.2, CP1.3, CP1.4, CP1.5, CP1.6, CP1.7, CP1.8, CP1.9, CP1.10)

return(values1)

}


set.seed(124)

imp.par <- 2

count1 <- replicate(100, situation2(1000, imp.par))

count1.bic <- matrix(0,11,100)
count1.aic <- matrix(0,11,100)
count1.cp <- matrix(0,11,100)
for(i in 1:100){
  for(j in 1:11){
    count1.bic[j,i] <- count1[j,i]
  }
  for(j in 12:22){
    count1.aic[j-11,i] <- count1[j,i]
  }
  for(j in 23:33){
    count1.cp[j-22,i] <- count1[j,i]
  }
}

best.model.bic <- 0
best.model.aic <- 0
best.model.cp <- 0
for(i in 1:100){
  best.model.bic[i] <- which.min(count1.bic[,i]) #BIC
  best.model.aic[i] <- which.min(count1.aic[,i]) #AIC
  best.model.cp[i] <- which.min(count1.cp[,i]) #marginal F test
}

count.bic <- 0
count.aic <- 0
count.cp <- 0

for(i in 1:100){
  if(best.model.bic[i]==(imp.par+1)){
    count.bic <- count.bic + 1
  }
  if(best.model.aic[i]==(imp.par+1)){
    count.aic <- count.aic + 1
  }
  if(best.model.cp[i]==(imp.par+1)){
    count.cp <- count.cp + 1
  }
}

count.bic #0
count.aic #10
count.cp #10

```

For the third situation I changed the variance of the X matrix.

```{r}

situation3 <- function(n, imp.par){

par <- 10
# coefficients
beta <- c(rep(1,imp.par),rep(0,par-imp.par))

# variables
X <- matrix(rnorm(n*par, 0, 0.001),n,par)
colnames(X) <- paste0("var", 1:par)

# outcome variable
y <- X %*% t(t(beta)) + rnorm(n)

dat <- data.frame(y,X)

# let's see the model
model1.0 <- lm(y ~ 1, data=dat)
model1.1 <- lm(y ~ var1, data=dat)
model1.2 <- lm(y ~ var1 + var2, data=dat)
model1.3 <- lm(y ~ var1 + var2 + var3, data=dat)
model1.4 <- lm(y ~ var1 + var2 + var3 + var4, data=dat)
model1.5 <- lm(y ~ var1 + var2 + var3 + var4 + var5, data=dat)
model1.6 <- lm(y ~ var1 + var2 + var3 + var4 + var5 + var6, data=dat)
model1.7 <- lm(y ~ var1 + var2 + var3 + var4 + var5 + var6 + var7, data=dat)
model1.8 <- lm(y ~ var1 + var2 + var3 + var4 + var5 + var6 + var7 + var8, data=dat)
model1.9 <- lm(y ~ var1 + var2 + var3 + var4 + var5 + var6 + var7 + var8 + var9, data=dat)
model1.10 <- lm(y ~ var1 + var2 + var3 + var4 + var5 + var6 + var7 + var8 + var9 + var10, data=dat)

BIC1.0 <- BIC(model1.0)
BIC1.1 <- BIC(model1.1)
BIC1.2 <- BIC(model1.2)
BIC1.3 <- BIC(model1.3)
BIC1.4 <- BIC(model1.4)
BIC1.5 <- BIC(model1.5)
BIC1.6 <- BIC(model1.6)
BIC1.7 <- BIC(model1.7)
BIC1.8 <- BIC(model1.8)
BIC1.9 <- BIC(model1.9)
BIC1.10 <- BIC(model1.10)

AIC1.0 <- AIC(model1.0)
AIC1.1 <- AIC(model1.1)
AIC1.2 <- AIC(model1.2)
AIC1.3 <- AIC(model1.3)
AIC1.4 <- AIC(model1.4)
AIC1.5 <- AIC(model1.5)
AIC1.6 <- AIC(model1.6)
AIC1.7 <- AIC(model1.7)
AIC1.8 <- AIC(model1.8)
AIC1.9 <- AIC(model1.9)
AIC1.10 <- AIC(model1.10)

CP1.0 <- sum(model1.0$residuals^2)/(summary(model1.10)$sigma^2) + 2 * model1.0$rank -
+ 100
CP1.1 <- sum(model1.1$residuals^2)/(summary(model1.10)$sigma^2) + 2 * model1.1$rank -
+ 100
CP1.2 <- sum(model1.2$residuals^2)/(summary(model1.10)$sigma^2) + 2 * model1.2$rank -
+ 100
CP1.3 <- sum(model1.3$residuals^2)/(summary(model1.10)$sigma^2) + 2 * model1.3$rank -
+ 100
CP1.4 <- sum(model1.4$residuals^2)/(summary(model1.10)$sigma^2) + 2 * model1.4$rank -
+ 100
CP1.5 <- sum(model1.5$residuals^2)/(summary(model1.10)$sigma^2) + 2 * model1.5$rank -
+ 100
CP1.6 <- sum(model1.6$residuals^2)/(summary(model1.10)$sigma^2) + 2 * model1.6$rank -
+ 100
CP1.7 <- sum(model1.7$residuals^2)/(summary(model1.10)$sigma^2) + 2 * model1.7$rank -
+ 100
CP1.8 <- sum(model1.8$residuals^2)/(summary(model1.10)$sigma^2) + 2 * model1.8$rank -
+ 100
CP1.9 <- sum(model1.9$residuals^2)/(summary(model1.10)$sigma^2) + 2 * model1.9$rank -
+ 100
CP1.10 <- sum(model1.10$residuals^2)/(summary(model1.10)$sigma^2) + 2 * model1.10$rank -
+ 100

values1 <- c(BIC1.0, BIC1.1, BIC1.2, BIC1.3, BIC1.4, BIC1.5, BIC1.6, BIC1.7, BIC1.8, BIC1.9, BIC1.10, AIC1.0, AIC1.1, AIC1.2, AIC1.3, AIC1.4, AIC1.5, AIC1.6, AIC1.7, AIC1.8, AIC1.9, AIC1.10, CP1.0, CP1.1, CP1.2, CP1.3, CP1.4, CP1.5, CP1.6, CP1.7, CP1.8, CP1.9, CP1.10)

return(values1)

}

set.seed(124)

imp.par <- 5

count1 <- replicate(100, situation3(100, imp.par))

count1.bic <- matrix(0,11,100)
count1.aic <- matrix(0,11,100)
count1.cp <- matrix(0,11,100)
for(i in 1:100){
  for(j in 1:11){
    count1.bic[j,i] <- count1[j,i]
  }
  for(j in 12:22){
    count1.aic[j-11,i] <- count1[j,i]
  }
  for(j in 23:33){
    count1.cp[j-22,i] <- count1[j,i]
  }
}

best.model.bic <- 0
best.model.aic <- 0
best.model.cp <- 0
for(i in 1:100){
  best.model.bic[i] <- which.min(count1.bic[,i]) #BIC
  best.model.aic[i] <- which.min(count1.aic[,i]) #AIC
  best.model.cp[i] <- which.min(count1.cp[,i]) #marginal F test
}

count.bic <- 0
count.aic <- 0
count.cp <- 0

for(i in 1:100){
  if(best.model.bic[i]==(imp.par+1)){
    count.bic <- count.bic + 1
  }
  if(best.model.aic[i]==(imp.par+1)){
    count.aic <- count.aic + 1
  }
  if(best.model.cp[i]==(imp.par+1)){
    count.cp <- count.cp + 1
  }
}

count.bic #0
count.aic #6
count.cp #7

```






