---
title: "Praca domowa 6"
author: "Piotr Obarski"
date: "Modele liniowe i mieszane"
output: 
  html_document:
  toc: TRUE
---

Strategy is to make up some data, check which criteria chooses the right model most often and then modify it.

Firstly I will make up some variables, and also creat table with comibnations and check which combinantion corresponds to my right model.

```{r}
library(e1071)
n<-100
  
    X<- cbind(2+rnorm(n), runif(n), sample(c(0,1), size=n, replace=TRUE, prob=c(0.5, 0.5)), sample(c(0.5,3,2,5,1), size=n, replace=TRUE), sample(c(0,1), size=n, replace=TRUE, prob=c(0.25, 0.75)), log(abs(rnorm(n))), rexp(n), rpois(n, 5), rcauchy(n), sample(c(3,1), size=n, replace=TRUE, prob=c(0.25, 0.75)))

cof<-c(1,5,0.5,0.33,2)
y<- cof[1]*X[,1]*cof[2]*X[,2]*cof[3]*X[,3]*cof[4]*X[,4]*cof[5]*X[,5]+rnorm(100)
X<-data.frame(X)
summary(lm(y~ ., data=X))
model <- lm(y ~ ., data=X)

AIC(model)
BIC(model)


comb <- bincombinations(10)[-1,]
crit <- matrix(0, nrow(comb), 2)

a<-intersect(which(comb[,1]==1),which(comb[,2]==1))
b<-intersect(which(comb[,3]==1),which(comb[,4]==1))
c<-which(comb[,5]==1)
d<-intersect(a,b)
intersect(d,c)
comb[992,]
```

# Scenario I

```{r}
n<-100

set.seed(1612)
outcome<-matrix(0, n ,3)
outcome<-replicate(100, {
  
    X<- cbind(2+rnorm(n), runif(n), sample(c(0,1), size=n, replace=TRUE, prob=c(0.5, 0.5)), sample(c(0.5,3,2,5,1), size=n, replace=TRUE), sample(c(0,1), size=n, replace=TRUE, prob=c(0.25, 0.75)), log(abs(rnorm(n))), rexp(n), rpois(n, 5), rcauchy(n), sample(c(3,1), size=n, replace=TRUE, prob=c(0.25, 0.75)))
colnames(X) <- c("X1", "X2", "X3", "X4", "X5", "X6", "X7", "X8", "X9", "X10")
    
    
cof<-c(1,5,0.5,0.33,2)
y<- cof[1]*X[,1]*cof[2]*X[,2]*cof[3]*X[,3]*cof[4]*X[,4]*cof[5]*X[,5]+rnorm(n)
X<-data.frame(X)
  
  
  for (i in 1:nrow(comb)) {
  form <- paste0("y~",
                 paste0("X",which(comb[i,]==1), collapse="+"))
  model <- lm(as.formula(form), data=X)
  crit[i,1] <- AIC(model)
  crit[i,2] <- BIC(model)
}
a<-(summary(lm(y~. , data=X))[[4]][2:11,4] < 0.05)==c(rep(TRUE,5),rep(FALSE, 5))
a<-sum(a)
if(a==10) {
  a<-992 
  } else {
  a<-0
  }
colnames(crit) <- c("AIC", "BIC")
crit <- data.frame(crit)

c(which(crit[,1]==min(crit[,1])), which(crit[,2]==min(crit[,2])), a) 
})
outcome<-t(outcome)
outcome<-(outcome==992)
apply(outcome,2,mean)

```

We see that BIC chooses the right model most often.

# Scenario II

Let's do the same thing, but change the seed.

```{r}

n<-100
set.seed(200)
outcome<-matrix(0, n ,3)
outcome<-replicate(100, {
  
    X<- cbind(2+rnorm(n), runif(n), sample(c(0,1), size=n, replace=TRUE, prob=c(0.5, 0.5)), sample(c(0.5,3,2,5,1), size=n, replace=TRUE), sample(c(0,1), size=n, replace=TRUE, prob=c(0.25, 0.75)), log(abs(rnorm(n))), rexp(n), rpois(n, 5), rcauchy(n), sample(c(3,1), size=n, replace=TRUE, prob=c(0.25, 0.75)))
colnames(X) <- c("X1", "X2", "X3", "X4", "X5", "X6", "X7", "X8", "X9", "X10")
    
    
cof<-c(1,5,0.5,0.33,2)
y<- cof[1]*X[,1]*cof[2]*X[,2]*cof[3]*X[,3]*cof[4]*X[,4]*cof[5]*X[,5]+rnorm(n)
X<-data.frame(X)
  
  
  for (i in 1:nrow(comb)) {
  form <- paste0("y~",
                 paste0("X",which(comb[i,]==1), collapse="+"))
  model <- lm(as.formula(form), data=X)
  crit[i,1] <- AIC(model)
  crit[i,2] <- BIC(model)
}
a<-(summary(lm(y~. , data=X))[[4]][2:11,4] < 0.05)==c(rep(TRUE,5),rep(FALSE, 5))
a<-sum(a)
if(a==10) {
  a<-992 
  } else {
  a<-0
  }
colnames(crit) <- c("AIC", "BIC")
crit <- data.frame(crit)

c(which(crit[,1]==min(crit[,1])), which(crit[,2]==min(crit[,2])), a) 
})
outcome<-t(outcome)
outcome<-(outcome==992)
apply(outcome,2,mean)


```

So we see that in this case F-test chooses the right model most often.


# Scenario III

Let's do the same thing, but change the seed.

```{r}

n<-100
set.seed(201)
outcome<-matrix(0, n ,3)
outcome<-replicate(100, {
  
    X<- cbind(2+rnorm(n), runif(n), sample(c(0,1), size=n, replace=TRUE, prob=c(0.5, 0.5)), sample(c(0.5,3,2,5,1), size=n, replace=TRUE), sample(c(0,1), size=n, replace=TRUE, prob=c(0.25, 0.75)), log(abs(rnorm(n))), rexp(n), rpois(n, 5), rcauchy(n), sample(c(3,1), size=n, replace=TRUE, prob=c(0.25, 0.75)))
colnames(X) <- c("X1", "X2", "X3", "X4", "X5", "X6", "X7", "X8", "X9", "X10")
    
    
cof<-c(1,5,0.5,0.33,2)
y<- cof[1]*X[,1]*cof[2]*X[,2]*cof[3]*X[,3]*cof[4]*X[,4]*cof[5]*X[,5]+rnorm(n)
X<-data.frame(X)
  
  
  for (i in 1:nrow(comb)) {
  form <- paste0("y~",
                 paste0("X",which(comb[i,]==1), collapse="+"))
  model <- lm(as.formula(form), data=X)
  crit[i,1] <- AIC(model)
  crit[i,2] <- BIC(model)
}
a<-(summary(lm(y~. , data=X))[[4]][2:11,4] < 0.05)==c(rep(TRUE,5),rep(FALSE, 5))
a<-sum(a)
if(a==10) {
  a<-992 
  } else {
  a<-0
  }
colnames(crit) <- c("AIC", "BIC")
crit <- data.frame(crit)

c(which(crit[,1]==min(crit[,1])), which(crit[,2]==min(crit[,2])), a) 
})
outcome<-t(outcome)
outcome<-(outcome==992)
apply(outcome,2,mean)

```

We see that AIC chooses the right model most often.

# Wniosek

Warto mieć dobre ziarno.