---
title: "Untitled"
author: "Emilia Pompe"
date: "Thursday, November 26, 2015"
output: pdf_document
---
## Emilia Pompe
## Homework 8

I performed 3 different types of regression on the Drosophila dataset.
```{r, warning=FALSE, message=FALSE}
library(PBImisc)
library(lasso2)
library(ggplot2)
library(reshape2)

head(Drosophila$bs)
dataset <- Drosophila$bs[,1:42]
```

### Scenario a). 

Standard linear regression is performed
```{r}
summary(lm(pc1~., data=dataset))

```
Let us take a look at the correlation structure in this dataset:
```{r, fig.width=4, fig.height=4}
corr = cor(dataset)
corr[1:6,1:6]

heatmap(1-abs(corr),symm=T,Rowv=NA,Colv=NA)
```

Some predictors seem to be quite strongly correlated hence standard mle's may have large variance.

### Scenario b).
Below I performed ridge regression. It can be observed from the plot that the larger the values of lambda, the closer are the parameters to 0. However, they do not actually give values equal to 0 - that is why ridge regression helps to shrink variance of estimates, bot does not perform model selection. 
```{r, fig.width=7, fig.height=4}
lambdas <- 10^seq(from= -4, by=0.1, to=10)
coefs <- sapply(lambdas, function(lambda) {
  lm.ridge(pc1~., data=dataset, lambda=lambda)$coef
})
df <- data.frame(t(coefs), lambdas)
df.long <- melt(df, id=c("lambdas"))
p <- ggplot(df.long, aes(x=lambdas, y=value, color=variable, 
                         group=factor(variable))) +geom_line(size=1) + theme(legend.position = "none") + 
                         scale_x_log10()
p
```

### Scenario c).

In the last scenario I used LASSo regression. For small values of lambda some parameters were estimated as 0, which means that LASSO did also a model selection, indicating which variables were most significant. The larger the parameters, the further are our parameters from 0.

```{r, fig.width=7, fig.height=4,warning=FALSE}
lambdas <- 10^seq(from= -4, by=0.1, to=10)
lambdas <- seq(from= 0.0, by=0.001, to=0.05)
coefs <- sapply(lambdas, function(lambda) {
  l1ce(pc1~., data=dataset, bound=lambda, absolute.t=TRUE)$coef
})
df2 <- data.frame(t(coefs), lambdas)[,-1] # I did not want to show a plot for intercept
df2.long <- melt(df2, id=c("lambdas"))

p <- ggplot(df2.long, aes(x=lambdas, y=value, color=variable, 
                         group=factor(variable))) +geom_line(size=1) + theme(legend.position = "none") + 
  scale_x_log10()
p

```