---
title: "PracaDom8"
author: "Magdalena Bogdańska"
date: "Modele liniowe i mieszane"
output: 
  html_document:
    toc: TRUE
---
# Task

For the Drosophila study check what will happen with coefficients if:

a) all variables are included into a model and standard MLE estimates are calculated, b) ridge regression is applied, c) lasso regression is applied.

For points b) and c) present how model coefficient behaves as a function of penalty/parameter.

```{r, warning=FALSE, message=FALSE}
library(PBImisc)
attach(Drosophila)
names(Drosophila)
head(bs)
df<-bs[,1:42]
```
First 41 variables are genotypes for selected loci, last 5 are phenotypes.
We'll consider a Drosophila study with 41 loci and a single quantitative trait (shape of proboscis) denoted by variable pc1.

# Standard MLE estimates
```{r, warning=FALSE, message=FALSE}
(coef1<-lm(pc1~., data=df)$coefficients)
range(coef1)
```
All coefficients are relatively small.

# Ridge regression
```{r, warning=FALSE, message=FALSE}
lambdas2 <- 10^seq(0,6.5,0.1)
coef2 <- sapply(lambdas2, function(par) {
  lm.ridge(pc1~., data=df, lambda=par)$coef
})
df2 <- data.frame(t(coef2), lambdas2)

library(tidyr)
df3<-gather(df2[,colnames(df2)!="X.Intercept."], variables, values, -lambdas2)

library(ggplot2)
ggplot(df3, aes(x=lambdas2, y=values, color=variables)) +
  geom_line() + ylab("coefficients") +   scale_x_log10() + theme(legend.position="none")
```

# Lasso regression
```{r, warning=FALSE, message=FALSE}
library(lasso2)
lambdas3 <- 10^seq(-5,1,0.1)
coef3 <- sapply(lambdas3, function(lambda) {
  l1ce(pc1~., data=df, bound=lambda, absolute.t=TRUE)$coef
})

df4 <- data.frame(t(coef3), lambdas3)
df5<-gather(df4[,colnames(df4)!="X.Intercept."], variables, values, -lambdas3)

ggplot(df5, aes(x=lambdas3, y=values, color=variables)) +
  geom_line() + ylab("coefficients") +   scale_x_log10() + theme(legend.position="none")
```
