---
title: "Praca domowa 2"
author: "Katarzyna Kanska"
date: "Modele liniowe i mieszane"
output: 
  html_document:
    toc: TRUE
---

## Problem
Compare range distribution (or studentized range distribution) for SNK/HSD tests between two scenarios:

* all groups have equal size (so called balanced case),
* some groups have unequal size (so called unbalanced case).
You may use KS-test or QQplots to verify if both distributions are equal.

Try different number of groups and different ways in which classes may be unequal.

## Solution

### Balanced case - dependence on number of groups

Fistly, let's set some paramethers:

```{r, warning=FALSE, message=FALSE}
# size of (whole) sample
N <- 10000
# number of groups
k <- c(10, 50, 100)
# number of obserwations in each group
m <- N/k
# number of simulations = size of the range distribution sample
S <- 5000
```

Now we will compare the range distribution for three scenarios - where we have 10, 50 and 100 groups.

```{r, warning=FALSE, message=FALSE}
# here we store our samples from range distributions
samples <- matrix(0, nrow=S, ncol=3)

set.seed(7)

for (i in 1:3){
  
  CalcDiff <- function(k, m, groups) {
    x <- rnorm(k*m)
    means <- tapply(x, groups, mean)
    difference <- max(means) - min(means)
    return(difference)
  }
  
  groups <- rep(1:k[i], times=m[i])
  
  samples[,i] <- replicate(S, CalcDiff(k[i],m[i],groups))

}

```

For the graphical analysis we will use histograms.

```{r, warning=FALSE, message=FALSE}
par(mfrow=c(3,1))
hist(samples[,1])
hist(samples[,2])
hist(samples[,3])

```

We see that the distributions are **right-skewed**. If there are more groups, the ranges are bigger (the histogram ''is moving towards right'' and ''becomes wider'').

Now we perform a formal test for comparing distibutions - Kolmogorov-Smirnov Test.
```{r, warning=FALSE, message=FALSE}

ks.test(samples[,1], samples[,2])
ks.test(samples[,2], samples[,3])
ks.test(samples[,3], samples[,1])

```

In every case we have to reject the null hypothesis that the samples are drawn from the same (continuous) distribution.

### Unbalanced case - dependence on type of unequality of groups

In this section we set the number of groups to 10 and see how a type of groups' unequality affects the range distribution.

```{r, warning=FALSE, message=FALSE}
# number of groups
k <- 10

# here we store our samples from range distributions
samples2 <- matrix(0, nrow=S, ncol=3)
```


Firstly, let's consider a case in which we have one big group and the rest of the gropus are equally large.

```{r, warning=FALSE, message=FALSE}
set.seed(7)

groups <- sample(1:k, size=N, replace=TRUE, prob = c(0.4, rep(0.6, times=(k-1))/(k-1)))

samples2[,1] <- replicate(S, CalcDiff(k,N/k,groups))
```


Then we consider a case in which we have two big groups and the rest of the gropus are equally large.

```{r, warning=FALSE, message=FALSE}
set.seed(7)

groups <- sample(1:k, size=N, replace=TRUE, prob = c(0.25, 0.25, rep(0.5, times=(k-2))/(k-2)))

samples2[,2] <- replicate(S, CalcDiff(k,N/k,groups))
```

Finally, we consider a case in which all the groups have unequal sizes.

```{r, warning=FALSE, message=FALSE}
set.seed(7)

groups <- sample(1:k, size=N, replace=TRUE, prob = (1:k)/sum(1:k))

samples2[,3] <- replicate(S, CalcDiff(k,N/k,groups))
```

For the graphical analysis we will use histograms.

```{r, warning=FALSE, message=FALSE}
par(mfrow=c(3,1))
hist(samples2[,1])
hist(samples2[,2])
hist(samples2[,3])

```

We see that the distributions are **right-skewed**. When there are one or two gropus with equal sizes and the rest of the groups are small, the distributions seem to be similar. When all the groups have unequal sizes, the distribution looks different - its tail is longer.

Now we perform a formal test for comparing distibutions - Kolmogorov-Smirnov Test.
```{r, warning=FALSE, message=FALSE}

ks.test(samples2[,1], samples2[,2])
ks.test(samples2[,2], samples2[,3])
ks.test(samples2[,3], samples2[,1])

```

These results confirms our graphical analysis. If there are one or two bigger groups and the other groups are small and equal in size, the distribustion is the same.

But in the case where every group had a different size, the distribution changed.

### Balanced vs unbalanced case

In this section we will compare the range distribution for 10 groups differing in size.

```{r, warning=FALSE, message=FALSE}
par(mfrow=c(4,1))
hist(samples[,1])
hist(samples2[,1])
hist(samples2[,2])
hist(samples2[,3])

```

We see that the 'more unequal' the groups' sizes are, the longer their tails are.

```{r, warning=FALSE, message=FALSE}

ks.test(samples[,1], samples2[,1])
ks.test(samples[,1], samples2[,2])
ks.test(samples[,1], samples2[,3])

```

In every case we have to reject the null hypothesis, which means that **the range distributions are different for balanced and unbalanced case**. This proves the importance of assumption in SNK/HSD tests that the groups have to be equal in size.
