---
title: "Homework 4"
author: "Katarzyna Kañska"
date: "26.10.2015"
output: html_document
---

We assume that residuals $\varepsilon \sim \mathcal{N}(0,1)$.
Recall that power of statistical test in defined by $\mathbb{P}(\text{reject }\mathrm{H_0} | \mathrm{H_{alternative}} \text{ is true})$.

### Create a scenario in which some main effects (main=additive) will be of the size 0.7. Use simulations to calculate the power of F test for main effects and power of test for interactions.

```{r, warning=FALSE, message=FALSE}
set.seed(77)

# size of (whole) sample
N <- 900
# number of groups (variable 1)
k1 <- 2
# number of groups (variable 2)
k2 <- 3
# number of obserwations in each group
m1 <- N/k1
m2 <- N/k2

# only residuals
Y <- rnorm(N)

# groups
x1 <- rep(c("I", "II"), times=m1)
x2 <- rep(c("A", "B", "C"), times=m2)

# differing means
Y[x1=="II"] <- Y[x1=="II"] + 2

Y[x2=="A"] <- Y[x2=="A"] + 0.1
Y[x2=="B"] <- Y[x2=="B"] + 0.3
Y[x2=="C"] <- Y[x2=="C"] + 0.85

myData <- data.frame(Y, x1, x2)

```

```{r, warning=FALSE, message=FALSE}
library("ggplot2")
ggplot(myData, aes(x=x1, y=Y)) + geom_boxplot()
ggplot(myData, aes(x=x2, y=Y)) + geom_boxplot()

```

Boxplots indicate diffrences in means between groups.

```{r, warning=FALSE, message=FALSE}
par(mfrow=c(1,2))
interaction.plot(myData$x1, myData$x2, myData$Y)
interaction.plot(myData$x2, myData$x1, myData$Y)
```

Lines are parallel, so there is no evidence of interaction between x1 and x2.

```{r, warning=FALSE, message=FALSE}
model1 <- lm(Y ~ x1 + x2, data = myData)
summary(model1)
```

The main effect for group C (variable x2) is of the size 0.7 (as the standard deviation equals 1 in our case).

```{r, warning=FALSE, message=FALSE}
model2 <- lm(Y ~ x1 * x2, data = myData)
summary(model2)
```

We see that interactions are not statistically significant. Therefore to calculate the power of F test for main effects we use the model without interactions.

```{r, warning=FALSE, message=FALSE}
set.seed(777)

M <- 100
pvals <- replicate(M, {

  Y <- rnorm(N)
  
  # differing means
  Y[x1=="II"] <- Y[x1=="II"] + 2
  
  Y[x2=="A"] <- Y[x2=="A"] + 0.1
  Y[x2=="B"] <- Y[x2=="B"] + 0.3
  Y[x2=="C"] <- Y[x2=="C"] + 0.85

  myData <- data.frame(Y, x1=factor(x1), x2=factor(x2))
  
  tmp <- summary(aov(Y~x1+x2, data=myData))
  c(tmp[[1]][1,5], tmp[[1]][2,5])
})
    
df <- data.frame(p1=pvals[1,], p2=pvals[2,])
    
ggplot(df, aes(x=p1)) + geom_histogram() + xlim(0,1) +
      ggtitle(paste0("Distribution of p-values for significance of variable x1\n\n",
                     round(100*mean(pvals[1,] < 0.05),1), "% of p-values smaller than 0.05 \n\n", 
                     round(100*mean(pvals[1,] >= 0.05),1), "% of p-values larger than 0.05"))

ggplot(df, aes(x=p2)) + geom_histogram() + xlim(0,1) +
      ggtitle(paste0("Distribution of p-values for significance of variable x2\n\n",
                     round(100*mean(pvals[2,] < 0.05),1), "% of p-values smaller than 0.05 \n\n", 
                     round(100*mean(pvals[2,] >= 0.05),1), "% of p-values larger than 0.05"))
```

The power of F test for main effects is equal 100%.

In this example it is not possible to find the power of F test for interactions because $\mathrm{H_{alternative}}$ is not true ($\mathrm{H_0}$ is that there are no interactions which is true).


### Create a scenario in which some interaction effects will be of the size 0.7. Use simulations to calculate the power of F test for main effects and power of test for interactions (since there are no main effects test for them should be on the level alpha).

```{r, warning=FALSE, message=FALSE}
set.seed(77)

# only residuals
Y <- rnorm(N)

# adding interaction
Y[x1=="II" & x2=="B"] <- Y[x1=="II" & x2=="B"] + 0.9

myData2 <- data.frame(Y, x1, x2)

```

```{r, warning=FALSE, message=FALSE}
library("ggplot2")
ggplot(myData2, aes(x=x1, y=Y)) + geom_boxplot()
ggplot(myData2, aes(x=x2, y=Y)) + geom_boxplot()

```

Boxplots do not show much diffrence in means between groups.

```{r, warning=FALSE, message=FALSE}
par(mfrow=c(1,2))
interaction.plot(myData2$x1, myData2$x2, myData2$Y)
interaction.plot(myData2$x2, myData2$x1, myData2$Y)
```

Lines are not parallel, so there is clear evidence of interaction between x1 and x2.

```{r, warning=FALSE, message=FALSE}
model1 <- lm(Y ~ x1 + x2, data = myData2)
summary(model1)
```

We see that main effects are statistically significant when the model is not considering interactions.

```{r, warning=FALSE, message=FALSE}
model2 <- lm(Y ~ x1 * x2, data = myData2)
summary(model2)
```

But when we introduce the interactions, the main effects loose their significance. The interaction effect between groups II and B is of the size 0.7 (as the standard deviation equals 1 in our case).

```{r, warning=FALSE, message=FALSE}
set.seed(777)

M <- 100
pvals <- replicate(M, {

  Y <- rnorm(N)
  
  # adding interaction
  Y[x1=="II" & x2=="B"] <- Y[x1=="II" & x2=="B"] + 0.9

  myData2 <- data.frame(Y, x1=factor(x1), x2=factor(x2))
  
  summary(aov(Y~x1*x2, data=myData2))[[1]][3,5]
})
    
df <- data.frame(p=pvals)
    
ggplot(df, aes(x=p)) + geom_histogram() + xlim(0,1) +
      ggtitle(paste0("Distribution of p-values for significance of interactions\n\n",
                     round(100*mean(pvals < 0.05),1), "% of p-values smaller than 0.05 \n\n", 
                     round(100*mean(pvals >= 0.05),1), "% of p-values larger than 0.05"))
```

The power of F test for interactions is equal 100%.

In this example it is not possible to find the power of F test for main effects because $\mathrm{H_{alternative}}$ is not true.



### Create a scenario in which there are main effects and interaction effects of the size 0.7. Use simulations to calculate the power of F test for main effects and power of test for interactions.

```{r, warning=FALSE, message=FALSE}
set.seed(77)

# only residuals
Y <- rnorm(N)

# differing means
Y[x1=="II"] <- Y[x1=="II"] + 0.5

Y[x2=="C"] <- Y[x2=="C"] + 0.7

# adding interactions
Y[x1=="II" & x2=="B"] <- Y[x1=="II" & x2=="B"] + 0.9

myData3 <- data.frame(Y, x1, x2)

```

```{r, warning=FALSE, message=FALSE}
library("ggplot2")
ggplot(myData3, aes(x=x1, y=Y)) + geom_boxplot()
ggplot(myData3, aes(x=x2, y=Y)) + geom_boxplot()

```

From the boxplots we see differences in means between groups.

```{r, warning=FALSE, message=FALSE}
par(mfrow=c(1,2))
interaction.plot(myData3$x1, myData3$x2, myData3$Y)
interaction.plot(myData3$x2, myData3$x1, myData3$Y)
```

Lines are not parallel, so there is evidence of interaction between x1 and x2.

```{r, warning=FALSE, message=FALSE}
model1 <- lm(Y ~ x1 + x2, data = myData3)
summary(model1)
```

We see that main effects are statistically significant.

```{r, warning=FALSE, message=FALSE}
model2 <- lm(Y ~ x1 * x2, data = myData3)
summary(model2)
```

Also interactions are significant.

The main effect for group II is of the size 0.7, as well as the interaction effect between groups II and B.

```{r, warning=FALSE, message=FALSE}
set.seed(777)

M <- 100
pvals <- replicate(M, {

  Y <- rnorm(N)

  # differing means
  Y[x1=="II"] <- Y[x1=="II"] + 0.5
  
  Y[x2=="C"] <- Y[x2=="C"] + 0.7
  
  # adding interactions
  Y[x1=="II" & x2=="B"] <- Y[x1=="II" & x2=="B"] + 0.9
  
  myData3 <- data.frame(Y, x1=factor(x1), x2=factor(x2))
  
  tmp <- summary(aov(Y~x1*x2, data=myData3))
  c(tmp[[1]][1,5], tmp[[1]][2,5], tmp[[1]][3,5])
})
    
df <- data.frame(p1=pvals[1,], p2=pvals[2,], p3=pvals[3,])
    
ggplot(df, aes(x=p1)) + geom_histogram() + xlim(0,1) +
      ggtitle(paste0("Distribution of p-values for significance of variable x1\n\n",
                     round(100*mean(pvals[1,] < 0.05),1), "% of p-values smaller than 0.05 \n\n", 
                     round(100*mean(pvals[1,] >= 0.05),1), "% of p-values larger than 0.05"))

ggplot(df, aes(x=p2)) + geom_histogram() + xlim(0,1) +
      ggtitle(paste0("Distribution of p-values for significance of variable x2\n\n",
                     round(100*mean(pvals[2,] < 0.05),1), "% of p-values smaller than 0.05 \n\n", 
                     round(100*mean(pvals[2,] >= 0.05),1), "% of p-values larger than 0.05"))
    
ggplot(df, aes(x=p3)) + geom_histogram() + xlim(0,1) +
      ggtitle(paste0("Distribution of p-values for significance of interactions\n\n",
                     round(100*mean(pvals < 0.05),1), "% of p-values smaller than 0.05 \n\n", 
                     round(100*mean(pvals >= 0.05),1), "% of p-values larger than 0.05"))
```

The power of F test for main effects as well as for interactions is equal 100%.
