---
title: "One-way analysis of variance"
author: "Przemyslaw Biecek"
date: "Linear models with fixed and random effects"
output: 
  html_document:
    toc: TRUE
---

# Use Case 1: Leukaemia

## Intro

We are going to use data from Acute myeloid leukaemia AML study in order to practice the one-way ANOVA.

We observe an effect of analogue of D3 vitamin (2191) on patients with different genetic mutations.

The question: Are the mutations in genes that causes leukaemia related with the effect of D3 analogue (2191).


```{r, warning=FALSE, message=FALSE}
library(PBImisc)

head(AML)
# normalisation
AML$change.2191 = AML$CD14.2191 - AML$CD14.control
attach(AML)

summary(AML)
```

## Summaries and descriptive statistics

```{r, warning=FALSE, message=FALSE}
by(change.2191, Mutation, mean)

boxplot(change.2191~Mutation, data=AML)
abline(h=0, col='grey', lwd=2)
```

## One-way ANOVA

```{r, warning=FALSE, message=FALSE}
# with intercept
summary(aov(CD14.2191~Mutation))

(model.1.mut <- anova(lm(CD14.2191~Mutation)))

model.1.mut[1,5]

# without

summary(aov(change.2191~Mutation-1, data=AML))

(tmp <- summary(lm(change.2191~Mutation-1, data=AML)))
```

## Model matrix

```{r, warning=FALSE, message=FALSE}
head(model.matrix(CD14.2191~Mutation, data=AML))

head(model.matrix(change.2191~Mutation-1, data=AML))
```

## Model assumptions

```{r, warning=FALSE, message=FALSE}
library(car)
leveneTest(AML$CD14.2191, AML$Mutation)

shapiro.test(lm(CD14.2191~Mutation, data=AML)$residuals)
```

## Post-hoc tests

```{r, warning=FALSE, message=FALSE}
pairwise.t.test(CD14.2191, Mutation, data=AML)
```

# Use Case 2: Dose response studies

## Descriptive statistics

```{r, warning=FALSE, message=FALSE}
summary(vaccination)
attach(vaccination)
by(response, dose, mean)

library(lattice)
bwplot(response~dose, data=vaccination)
```

## One-way ANOVA

```{r, warning=FALSE, message=FALSE}
summary(aov(response~dose, data=vaccination))
```

## Post-hoc tests

```{r, warning=FALSE, message=FALSE}
model.va.3 = aov(response~dose, data=vaccination)
TukeyHSD(model.va.3)

par(mar=c(5,6,4,2))
plot(TukeyHSD(model.va.3), las=1)
```

## Tukey differences

```{r, warning=FALSE, message=FALSE}
library(agricolae)
model = aov(response~dose, vaccination)
HSD.test(model, "dose", console = TRUE)
```

## Student-Newman-Keuls  differences

```{r, warning=FALSE, message=FALSE}
SNK.test(model, "dose", console = TRUE)
```

## Least significant differences

```{r, warning=FALSE, message=FALSE}
LSD.test(model, "dose", p.adj="bonferroni", console = TRUE)
```

## Scheffe test

```{r, warning=FALSE, message=FALSE}
scheffe.test(model, "dose", console = TRUE)
```

## Model assumptions

Homogeneity of Variances

```{r, warning=FALSE, message=FALSE}
attach(vaccination)
bartlett.test(response~ dose, data=vaccination)

fligner.test(response~ dose, data=vaccination)

leveneTest(response, dose)

leveneTest(response, dose, center=mean)
```

# Contrasts

## List of contrasts

```{r, warning=FALSE, message=FALSE}
t(contr.treatment(5))

t(contr.helmert(5))

t(contr.poly(5))

t(contr.sdif(5))

model.matrix(response~dose, vaccination)[1:5,]

model.matrix(response~dose, vaccination,
 contrasts=list(dose=contr.helmert))[1:5,]

model.matrix(response~dose, vaccination,
 contrasts=list(dose=contr.poly))[1:5,]
```

## Contrasts in action

```{r, warning=FALSE, message=FALSE}
summary(lm(response~dose, data = vaccination))

kontr = contr.sdif(5)
summary(lm(response ~ dose, data = vaccination, contrasts=list(dose=kontr)))

model.va.7 <- lm(response~ordered(dose), data=vaccination)
summary(model.va.7)
```

## Compare all contrasts

```{r, warning=FALSE, message=FALSE}
options()$contrasts

contrs <- apropos("contr\\.")
coefs <- sapply(contrs, function(i)
 lm(response~dose, data = vaccination,
 contrasts=list(dose=get(i)))$coefficients
)
t(coefs)
```

# Home work

Finish all tasks in
https://rawgithub.com/pbiecek/LinearModels/master/MIMUW_2015/MaterialyPomocnicze/02_lab.html
and upload the knitr report (html+Rmd) on GitHub.

For each completed point (2-5) you will receive one point.

