---
title: "Analysis of covariance (ANCOVA)"
author: "Przemyslaw Biecek"
date: "Linear models with fixed and random effects"
output: 
  html_document:
    toc: TRUE
---

# Contrasts

This is a copy of the last section of lecture `02_oneway`.

Consider a categorical variable with some number of levels.

```{r, warning=FALSE, message=FALSE}
library(PBImisc)

head(vaccination)
summary(vaccination)

boxplot(response ~ dose, data=vaccination)
```

## List of contrasts

We have to code the categorical variable somehow. And the coding is important.

```{r, warning=FALSE, message=FALSE}

t(contr.treatment(5))

t(contr.helmert(5))

t(contr.poly(5))

t(contr.sdif(5))

model.matrix(response~dose, vaccination)[1:5,]

model.matrix(response~dose, vaccination,
 contrasts=list(dose=contr.helmert))[1:5,]

model.matrix(response~dose, vaccination,
 contrasts=list(dose=contr.poly))[1:5,]
```

## Contrasts in action

```{r, warning=FALSE, message=FALSE}
summary(lm(response~dose, data = vaccination))

kontr = contr.sdif(5)
summary(lm(response ~ dose, data = vaccination, contrasts=list(dose=kontr)))

model.va.7 <- lm(response~ordered(dose), data=vaccination)
summary(model.va.7)
```

## Compare all contrasts

```{r, warning=FALSE, message=FALSE}
options()$contrasts

contrs <- apropos("contr\\.")
coefs <- sapply(contrs, function(i)
 lm(response~dose, data = vaccination,
 contrasts=list(dose=get(i)))$coefficients
)
t(coefs)
```


## Your turn

In small groups, 2-3 people each, find an example application for each type of contrast.


# Order of tests


Let us consider a simple model with two variables `X` and `Z`. Note that they are correlated and both have some relation with variable `Y`.

```{r}
library(PBImisc)
head(YXZ)
cor(YXZ)
```

## What we are testing?

There is a lot of approaches to test the relation between X/Z and Y. Let's check them.

```{r}
# this is quite often
# type III - marginal
summary(lm(Y~X+Z, data=YXZ))

# is it a surprise?
# type I - sequential
anova(lm(Y~X+Z, data=YXZ))
anova(lm(Y~Z+X, data=YXZ))

# What about this?
anova(lm(Y~X, data=YXZ))
anova(lm(Y~Z, data=YXZ))

```

!(YXZ2.png)

## Your turn

In small groups, 2-3 people each, explain what has happen. 
Why this has happened?

Find example in which one would like to use type I test. And when type III test would be more useful?

# Nested or crossed?

Let's play with data about prices of apartments in Warsaw.

```{r}
library(PBImisc)
head(apartments)
```

Let's select 7 variables that look like factors.

```{r}
head(apartments[,c(1,2,5,6,7,10,14)])
```

## Your turn

In small groups, 2-3 peoples each, choose how to code each variable.
What kind of contrast should / might be used?

For pairs of variables like type/condition, district/street, year/month, discuss if they should be coded as nested or crossed variables?

## Try with R

What will be the difference between

```{r}
model1 <- lm(m2.price ~ factor(year)+factor(month), data=apartments)

model2 <- lm(m2.price ~ factor(year)/factor(month), data=apartments)

model3 <- lm(m2.price ~ factor(year)*factor(month), data=apartments)

summary(model1)
summary(model2)
summary(model3)

# What about: type/condition, district/street

```

# Home works

Let's see home works `agnieszka_sitko_05.html`, `hw5_emilia_pompe.html`, `Katarzyna_Kanska_h5.Rmd`.

https://rawgit.com/kgajewska/LinearModels/c9db0bc8cdd4cdd3c5187ee8ebea680a17292670/MIMUW_2015/HomeWorks/Fifth/karolina_gajewska_h5.html

## Home work for 12 XI 2015

Read the chapter 2.5 from ,,Analiza danych z programem R, Modele liniowe mieszane'' or in English (http://www.statisticshell.com/docs/ancova.pdf + how to do this in R http://www.stat.columbia.edu/~martin/W2024/R8.pdf).

Now consider scenario with ,,unequal slopes'' - an interaction between continuous and categorical variable. We are going to check when such interaction can be detected.

Create following scenarios:

Generate continuous variable `X_1` from distribution `U[0,1]`. Generate categorical variable `X_2` with 2 levels with equal frequency). Generate Y in a way that it depends on interaction of `X_1` and `X_2`.

1. Add random noise from N(0,1), all ANCOVA assumptions are met. 
2. Choose other distribution for epsilons, find a distribution with the same scale as N(0,1) for which you will get drop in power.
3. Choose other distribution for epsilons in a way that the test for residual will not keep I type error rate.

For all these scenarios, produce plot that will compare power as a function of sample size. Moreover on this plot is should be visible that power for scenario 2 is lower than for scenario 1. And that I type error rate is different for scenario 3 and scenario 1.



