---
title: "Random coefficients"
author: "Przemyslaw Biecek"
date: "Linear models with fixed and random effects"
output: 
  html_document:
    toc: TRUE
---

# Model with random coefficients

## The `kidney` dataset

MDRD stands for The Modification of Diet in Renal Disease. It is an estimate of Glomerular Filtration Rate (GFR). It may be used to assess the performance of a kidney - the larger the better.

In the `kidney` data we have profiles of Graft function after kidney transplantation. Measurements start at day 7 after transplantation and last up to 60 months (for some patient shorter).

```{r, message=FALSE, warning=FALSE}
library(lme4)
library(PBImisc)
head(kidney)
```

We need to transform this dataset from wide format to long format. Also called RCV (row-column-value) format.

```{r, message=FALSE, warning=FALSE}
lkidney = reshape(kidney, direction="long", varying=list(9:16), v.names=c("MDRD"), times=c(1/4,1,3,6,12,24,36,60))
head(lkidney[,5:11])
```

And as we will see, it is a good idea to use log(time) instead of time.

```{r, message=FALSE, warning=FALSE}
lkidney$ltime = log(lkidney$time)
lkidney$title = factor(with(lkidney, paste("wiek ", recipient.age,"/",donor.age, ", zgod. ", discrepancy.AB,"/",discrepancy.DR, sep="")))
```

## Plots

We have data, now we can plot it.
We are adding linear and quadratic trend to each group.

```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=10}
library(lattice)
# panel for lattice package
panel.with.square <- function(...){
  nx = list(...)$x
  ny = list(...)$y
  cc = lm(ny~nx+I(nx^2))$coef
  panel.xyplot(...)
  panel.curve(x^2*cc[3]+x*cc[2]+cc[1], from=min(nx), to=max(nx), col="red")
}
# the plot
xyplot(MDRD ~ log(time) | title, lkidney, type = c("g","p","r"),
  xlab = "log(time after graft)", ylab = "MDRD", pch=19,
  panel = panel.with.square, subset= id>20 & id < 51)
```

## Models

We know how the data looks like, now we can fit some models.

Let's start with models with intercept only, with intercept and slope and only with slope.

```{r, message=FALSE, warning=FALSE}
(model0 = lmer(MDRD~1 + (1|id), data=lkidney, REML=FALSE))
(model1 = lmer(MDRD~ltime + (ltime|id), data=lkidney, REML=FALSE))
(model1bis = lmer(MDRD~ltime + (ltime-1|id) + (1|id), data=lkidney))
```

How we can extract individual coefficients from the model?

```{r, message=FALSE, warning=FALSE}
fixef(model1)
head(ranef(model1)$id)

randomCoefs = t(t(ranef(model1)$id) + fixef(model1))
head(randomCoefs)
```

## Fixed effects

We can do similar modelling for fixed effects only. 

Let's compare the model with mixed effects to the one with fixed effects only.

```{r, message=FALSE, warning=FALSE}
model1fixed = lm(MDRD~ltime:factor(id) + factor(id) - 1, data=lkidney)
fixedCoefs = matrix(coef(model1fixed), ncol=2)
head(fixedCoefs)
```

```{r, eval=FALSE}
lmList(MDRD ~ time | id, lkidney)
## Call: lmList(formula = MDRD ~ time | id, data = lkidney) 
## Coefficients:
##     (Intercept)         time
## 1      63.09062  0.219859695
## 2      55.55200  0.165792565
## 3      33.66915  0.157798211
## 4      39.20332  0.984699102
## 5      62.64902  0.126592898
## 6      39.30236  0.313399820
## 7      40.23394  0.113381365
## 8      61.81949  0.298376739
## 9      49.41234  0.750096638
## 10     32.69759  0.216655569
```

## Regularisation  

We are going to extract individual coefficient from both models and then we are going to compare these coefficients on a single plot.

What is happening with these coefficients?
  
```{r, message=FALSE, warning=FALSE, fig.width=10, fig.height=10}
plot(fixedCoefs[,1], fixedCoefs[,2], cex=1, pch=21, lwd=3, col="red3", xlab=expression(a[i]), ylab=expression(b[i]))

abline(v=fixef(model1)[1], col="grey", lwd=2, lty=2)
abline(h=fixef(model1)[2], col="grey", lwd=2, lty=2)
arrows(fixedCoefs[,1], fixedCoefs[,2], randomCoefs[,1], randomCoefs[,2], angle=15, length=0.1, lwd=1)
points(randomCoefs[,1], randomCoefs[,2], cex=1, pch=7, lwd=2, col="blue3")
```

## Higher order polynomials

The sky is the limit.

Up to now we have tried only model with linear trend. But figures suggest that quadratic trend may improve the fit.

Let's create a model with quadratic coefficients as well.

Then we can use log likelihood ratio test to compare these models.

```{r, message=FALSE, warning=FALSE}
(model2 = lmer(MDRD~ltime + I(ltime^2) + (ltime+I(ltime^2)|id), data= lkidney, REML=FALSE))

anova(model2,model1,model0)
```

## Other fixed effects

The random effects are only to define covariance matrix and allow for estimated for fixed effects.

But in fact we may be more interested in fixed effects.

So the usual strategy is to first work on variance matrix (random effects) and then tune and test fixed effects.

Here we are going to decide what influence the prognosis of MDRD kidney function. The three most important variables are: donor.age, number of drugs and time after transplantation.

```{r, message=FALSE, warning=FALSE}
(fullModel = lmer(MDRD~donor.age+ recipient.age+ therapy+ diabetes+ bpl.drugs+ discrepancy.AB+ discrepancy.DR + ltime + I(ltime^2) + (ltime + I(ltime^2)|id), data=lkidney, REML=FALSE))

(finalModel = lmer(MDRD~donor.age+ bpl.drugs+ ltime + I(ltime^2) + (ltime+I(ltime^2)|id), data=lkidney, REML=FALSE))

```


# The homework [7 I 2016]

The data `sleepstudy`  (lme4) allow to measure the effect of sleep deprivation of reaction time.

*The average reaction time per day for subjects in a sleep deprivation study. On day 0 the subjects had their normal amount of sleep. Starting that night they were restricted to 3 hours of sleep per night. The observations represent the average reaction time on a series of tests given each day to each subject.*

With the use of the `sleepstudy`:

- Plot the relation between `Reaction` and `Days` for each subject independently.
- Build a model with random coefficients for linear trend between `Reaction` and `Days`.
- Extract individual slopes for the relation from the model with random effects. Present summary statistics of these slopes (1Q, median, 3Q).


