---
title: "Homework Week 1"
author: "Annanina Koster"
date: "7 oktober 2015"
output: html_document
---

- - -

Praca domowa

Na zbiorze danych heights{PBImisc} oszacuj zależność liniową pomiędzy wzrostem żony a męża.

Wyznacz obliczeniowo przedział ufności dla współczynnika nachylenia.

Raport w formacie html z przedziałem ufności wyślij w GitHubie na adres

https://github.com/pbiecek/ModeleLiniowe/tree/master/MIMUW_2015

- - -

{Homework

At dataset heights {PBImisc} estimate a linear relationship between the increase of his wife and husband.
  
Determine computationally confidence interval for the slope.
  
The report in html format with a confidence interval send a Github address
  
https://github.com/pbiecek/ModeleLiniowe/tree/master/MIMUW_2015}
  
- - -
  
First of all I open the dataset 'heights' from the PBImisc library and store it in a dataset named data.heights. Then I plot the data to see if there is a positive linear relation between the heights of Husbands and Wifes. As can be seen, there is, so the variables don't have to be adjusted. 

```{r}
library("PBImisc")
data.heights <- heights

plot(data.heights$Wife, data.heights$Husband)
library(ggplot2)
ggplot(data.heights, aes(x=Wife, y=Husband))+geom_point()
ggplot(data.heights, aes(x=Wife, y=Husband))  +
  geom_point() + 
  geom_smooth(method="lm", formula=y~x) +
  ggtitle("Height Wife vs Height Husband")
  
```

Then a linear model is fitted to the data.

```{r}

model.heights <- lm(Husband~Wife, data=data.heights)
model.heights$coefficients
summary(model.heights)
```

Using the bootstrap method, a 95% confidence interval for the slope is calculated.

```{r}

set.seed(12)
N <- 5000
nwsp.heights <- replicate(N, {
  ndf.heights <- data.frame(x = data.heights$Wife,
                    y = model.heights$fitted.values + sample(model.heights$residuals))
  model2.heights <- lm(y~x, data=ndf.heights)
  model2.heights$coefficients
})
df.heights <- data.frame(intercept = nwsp.heights[1,], slope = nwsp.heights[2,])
par(mfrow=c(1,1))
hist(df.heights$intercept, 100, col="grey")
hist(df.heights$slope, 100, col="grey")

quantiles.heights <- quantile(df.heights$slope, probs=c(0.025, 0.975))
quantiles.heights

```
The 95% confidence interval of the slope is from 0.6891461 to 0.9739710.
