---
title: "Project 1, Phase 2"
author: "Emilia Pompe, Ma³gorzata Pujszo"
date: "Wednesday, November 11, 2015"
output: html_document
---
###Main idea
Our approach was to find a list of genes that may be used to differentiate all types of genes from each other. In our report also interactions with sex were taken into consideration. 

###Table of contents
 * Changing the data
 * Assumptions and two-way ANOVA
 * Models without interactions
 * Models with interactions
 * Final conclusions

###Changing the data

We have loaded required libraries.
```{r message=FALSE, error=FALSE, warning=FALSE, echo=FALSE}
library(MASS)
library(dplyr)
library(plyr)
library(car)
library(agricolae)
library(nortest)
```

Merging datasets looks similarly as in Phase 1 of the Project 1 but now we don't delete any outliers. 
```{r, echo=FALSE}
load("expression.cb1.rda")
load("expression.cb2.rda")
load("clinical.cb.rda")

expression <-rbind(expression.cb1, expression.cb2)
colnames(expression)<-gsub(".","-", colnames(expression), fixed=TRUE)
trial <-t(expression[,-1])
trial<- as.data.frame(trial)
colnames(trial) <- expression[,1]
trial$sampleID <- rownames(trial)
dataset<-merge(trial, clinical.cb, by="sampleID")
```

As before we removed the cancer with only 12 observations.
```{r, echo=FALSE}
dataset <- subset(dataset, X_cohort != "TCGA Formalin Fixed Paraffin-Embedded Pilot Phase II")
dataset$X_cohort<-factor(dataset$X_cohort)
```

###Assumptions and two-way ANOVA

We would like to check whether interactions of gender and type of cancer influence genes' expressions. In order to do this we would like to perform two-way ANOVA. However, we need to check normality of residuals and homogeneity of variances in gropus. We have decided to do these test simultanously in one loop.

We are checking normality of residuals using Shapiro test (p-values for each gene are remembered in shapiro_coeeff1). For homogeneity of variances we are performing Levene test (p-values are stored in levene1). M.anova is a matrix containing p-values of two-way ANOVA and gene's number.
```{r, eval=FALSE}
N <- 16115
shapiro_coeff1 <- rep(NA, times=N)
levene1 <- rep(NA, times=N)
M.anova <- matrix(NA, nrow=N, ncol=4)
M.anova[,4] <- 1:N
```

For each gene we are checking if each group contains more than treshold (in this case 20) observations. If it doesn't we are removing those groups in order not to infer from such little information. Here is a function which does it.
```{r}
threshold <- 20
filtered.dataset <- function(i) {
  y <- dataset[,i]
  d <- data.frame(y=y[!(is.na(y))], x=dataset$X_cohort[!(is.na(y))], 
                  sex = dataset$gender[!(is.na(y))])
  d <- subset(d, sex %in% c("FEMALE", "MALE"))
  d$sex <- factor(d$sex)
  tab <- as.data.frame(table(d$x, d$sex))
  bad.female <-intersect(which(tab[,3]<=threshold), which(tab[,2]=="FEMALE"))
  bad.male <-intersect(which(tab[,3]<=threshold), which(tab[,2]=="MALE"))
  d <- filter(d, !(x %in% tab[bad.female, 1] & sex=="FEMALE")) 
  d <- filter(d, !(x %in% tab[bad.male, 1] & sex=="MALE"))
  d$x <- factor(d$x)
  d$sex <- factor(d$sex)
  d
  }
```

Here is the loop performing our tests. 
```{r, eval=FALSE}
for(i in 2:(N+1)){
  d <- filtered.dataset(i)
  model <- lm(y~ x * sex, data=d)
  shapiro_coeff1[i-1] <- shapiro.test(rstandard(model))$p.value
  levene1[i-1] <- leveneTest(d$y~d$x*d$sex)[1,3]
  M.anova[i-1,1:3] <- anova(model)[1:3,5]
}
```

Now, we would like to find reasonable threshold for Shapiro test's p-value. We created a data frame with sorted p-values. 
```{r, eval=FALSE}
shapiro.data <- data.frame(nr=1:N, shapiro = shapiro_coeff1)
shapiro.data.new <- arrange(shapiro.data, shapiro)
```

Let's see what is happening with gene with the lowest p-value.
```{r, echo=FALSE}
load("phase2.results1.RData")
shapiro.data.new <- phase2.results1$shapiro.data.new
```
```{r, fig.height=3, fig.width=6}
par(mfrow=c(1, 2), mar=c(4, 4, 4, 4))
d <- filtered.dataset(shapiro.data.new$nr[1]+1)
model <- lm(y ~ x * sex, data=d)
qqnorm(rstandard(model))
abline(0, 1, col="red", lwd=3)
hist(rstandard(model), breaks=100, main=paste0("Histogram of standardized residuals\n", colnames(dataset)[shapiro.data.new$nr[1]+1]), cex.main=1)
```

As we can see the tail is heavy so we cannot perform ANOVA test in this case. We plotted several more examples, not all of them are shown here. We noticed that gene named GOLT1B with p-value of order 10^(-34) is quite good whereas gene named LSM1 with p-value of order 10^(-35) has heavy right tail.
```{r, fig.height=3, fig.width=6}
par(mfrow=c(1, 2), mar=c(4, 4, 4, 4))
i <- 5823
d <- filtered.dataset(i)
model <- lm(y~ x * sex, data=d)
qqnorm(rstandard(model))
abline(0, 1, col="red", lwd=3)
hist(rstandard(model), breaks=100, main=paste0("Histogram of standardized residuals\n", colnames(dataset)[i]), cex.main=1)
i <- 8149
d <- filtered.dataset(i)
model <- lm(y~ x * sex, data=d)
qqnorm(rstandard(model))
abline(0, 1, col="red", lwd=3)
hist(rstandard(model), breaks=100, main=paste0("Histogram of standardized residuals\n", colnames(dataset)[i]), cex.main=1)
```

We decided to take 10^(-34) as a threshold for p-value for this test.
Similarly, we need to find reasonable threshold for Levene test.
```{r, eval=FALSE}
levene.data <- data.frame(nr=1:N, levene = levene1)
levene.data.new <- arrange(levene.data, levene)
```

We have chosen significance level equal to 10^(-42). For the first gene with p-value higher than this cutoff, we displayed the highest and the lowest variances in groups. 
```{r, echo=FALSE}
levene.data.new <- phase2.results1$levene.data.new
```
```{r}
i <- levene.data.new[min(which(levene.data.new[,2]>= 10^(-42))), 1]+1 
d <- filtered.dataset(i)
model <- lm(y~ x * sex, data=d)
d$res <- model$residuals
variances <- ddply(d, .(x, sex), summarize, var = var(res))
min(variances$var)
max(variances$var)
```

In comparison to this variable, the highest and lowest variances in groups of the variable with the highest p-value are the following.
```{r}
i <- levene.data.new[which.max(levene.data.new$levene),1]+1 
d <- filtered.dataset(i)
model <- lm(y~ x * sex, data=d)
d$res <- model$residuals
variances <- ddply(d, .(x, sex), summarize, var = var(res))
min(variances$var)
max(variances$var)
```

And the same results for the lowest p-value.
```{r}
i <- levene.data.new[1,1]+1 
d <- filtered.dataset(i)
model <- lm(y~ x * sex, data=d)
d$res <- model$residuals
variances <- ddply(d, .(x, sex), summarize, var = var(res))
min(variances$var)
max(variances$var)
```

All our significance levels look as follows: 
```{r}
p.value.shapiro.threshold <- 10^(-34)
p.value.levene.threshold <- 10^(-42)
cancer.p.value <- 0.05
interaction.p.value <- 0.05
```

###Models without interactions
Firstly, we chose genes for which the assumptions of two-way ANOVA hold. For them we used p-value adjustment using the Holm method. used.genes is a vector of numbers of genes for which the effect of cancer is significant but the effect of interaction is not significant.
```{r, eval = FALSE}
p.value.genes <- intersect(which(shapiro_coeff1 > p.value.shapiro.threshold), 
                           which(levene1 > p.value.levene.threshold))
M.anova.part <- M.anova[p.value.genes,] 
M.anova.used <- cbind(matrix(p.adjust(as.vector(M.anova.part[,1:3]), method= "holm"), ncol=3), M.anova.part[,4]) 
M.anova.used <- as.data.frame(M.anova.used)
colnames(M.anova.used) <- c("cancer.pvalues", "sex.pvalues", "interaction.pvalues", "nr")
used.genes <- M.anova.used$nr[intersect(which(M.anova.used[,3]> interaction.p.value), which(M.anova.used[,1]< cancer.p.value))] 
```

For each of the chosen genes we created a linear model without interactions with sex. Afterwards, we performed the LSD test - it was chosen as the groups sizes are unbalanced. For each result of the LSD test we checked whether the gene with the highest mean (or with the lowest mean, respectively) is in its own group. We decided to do this, as these genes will have an easy interpretation. Finally, for each type of cancer we found many genes that may be used for detecting it. The most interesting ones are those for which difference of the means for the 'highest/lowest' type of cancer and the 'next one' is large. Here we used 'relative difference' - we divided difference in means by standard deviation of the variable.
```{r, eval = FALSE}
results <- data.frame(highest=rep(NA,times=N), lowest=rep(NA,times=N), 
                      difference.highest=rep(NA, times=N), difference.lowest=rep(NA, times=N))
for (k in used.genes){
  i <- k+1
  d<-filtered.dataset(i)
  model <- LSD.test(lm(y~ x, data=d), "x") 
  groups <- as.character(model$group[,"M"])
  high.mean <- grep("a", groups) 
  if(length(high.mean)==1){
    results$highest[k] <- as.character(model$group[,"trt"])[1]
    results$difference.highest[k] <- (model$groups[1, c("means")] - model$groups[2, c("means")])/sd(d$y)
  }
  low.mean <- grep(groups[length(groups)], groups) 
  if(length(low.mean)==1){
    results$lowest[k] <- as.character(model$group[,"trt"])[length(groups)] 
    results$difference.lowest[k] <- (model$groups[length(groups),c("means")] - model$groups[(length(groups)-1), c("means")])/sd(d$y)
  }
}
```

```{r, echo = FALSE}
load("phase2.results2.RData")
```

For each type of cancer we would like to present one gene which differentiates it from other ones. We chose genes with the highest/lowest 'relative differences', obtaining a data frame all.means.genes that looks as follows:
```{r}
highest.means.genes <- data.frame(name=rep(NA,times=nlevels(dataset$X_cohort)),
                                  difference=rep(NA,times=nlevels(dataset$X_cohort)),
                                  nr=rep(NA,times=nlevels(dataset$X_cohort)))
lowest.means.genes <- data.frame(name=rep(NA,times=nlevels(dataset$X_cohort)),
                                 difference=rep(NA,times=nlevels(dataset$X_cohort)),
                                 nr=rep(NA,times=nlevels(dataset$X_cohort)))
i <-1
for (cancer.names in levels(dataset$X_cohort)){
  v <- grep(cancer.names, results$highest)
  cancer.genes <- results[v,c(1,3,5)]
  highest.means.genes[i,] <- cancer.genes[which.max(cancer.genes$difference.highest),]
  v <- grep(cancer.names, results$lowest)
  cancer.genes <- results[v,c(2,4,5)]
  lowest.means.genes[i,] <- cancer.genes[which.min(cancer.genes$difference.lowest),]
  i <- i+1
}
all.means.genes <- cbind(highest.means.genes, lowest.means.genes)[,-4]
all.means.genes$name <- levels(dataset$X_cohort)
all.means.genes
```

We plotted 'plot.design' for each type of cancer (gene names are presented in the plot titles).
```{r, fig.width=8, fig.height=24}
par(mfrow=c(6,2), mar=c(2,2,2,2))
i <-1
for (cancer.names in levels(dataset$X_cohort)) {
 
  nr.gene <- ifelse(abs(all.means.genes[i,2])>=abs(all.means.genes[i,2]), all.means.genes[i,3], all.means.genes[i,5])
  d <- filtered.dataset(nr.gene+1)
  plot.design(y~x, data=d, main=paste0(cancer.names,", ",
                                       colnames(dataset)[nr.gene+1]), cex=0.7)
  if(cancer.names %in% c("TCGA Breast Cancer", "TCGA Acute Myeloid Leukemia")){
    y <- dataset[,nr.gene+1]
    d <- data.frame(y=y[!(is.na(y))], x=dataset$X_cohort[!(is.na(y))], sex = dataset$gender[!(is.na(y))])
    d <- subset(d, sex %in% c("FEMALE", "MALE"))
    bad <- names(which(table(d$x) < threshold)) 
    d <- filter(d, !(x %in% bad)) 
    d$x <- factor(d$x) 
    d$sex <- factor(d$sex)
    print(cancer.names)
    print(table(d$x, d$sex))
    }
  i <- i+1
}
```

One thing that needs to be clarified is the fact that our procedure might have deleted some significant levels e.g. all observations belonging to the Bladder Cancer. We presented a 'good' and a 'bad' example above.

For 'bads' we need to find the better examples of genes.
```{r, fig.width=8, fig.height=8}
  par(mfrow=c(2,2), mar=c(2,2,2,2))
  cancer.names <- "TCGA Breast Cancer"
  nr.gene <- 4825
  d <- filtered.dataset(nr.gene+1)
  plot.design(y~x, data=d, main=paste0(cancer.names,", ",
                                       colnames(dataset)[nr.gene+1]), cex=0.7)
  cancer.names <- "TCGA Bladder Cancer"
  nr.gene <- 2938
  d <- filtered.dataset(nr.gene+1)
  plot.design(y~x, data=d, main=paste0(cancer.names,", ",
                                       colnames(dataset)[nr.gene+1]), cex=0.7)
  cancer.names <- "TCGA Colon Cancer" 
  nr.gene <- 15920
  d <- filtered.dataset(nr.gene+1)
  plot.design(y~x, data=d, main=paste0(cancer.names,", ",
                                       colnames(dataset)[nr.gene+1]), cex=0.7)
  cancer.names <- "TCGA Lung Adenocarcinoma"
  nr.gene <- 10879
  d <- filtered.dataset(nr.gene+1)
  plot.design(y~x, data=d, main=paste0(cancer.names,", ",
                                       colnames(dataset)[nr.gene+1]), cex=0.7)

```

###Models with interactions

```{r, echo=FALSE}
load("phase2.results1.RData")
M.anova <- phase2.results1$M.anova
shapiro_coeff1 <- arrange(phase2.results1$shapiro.data.new, nr)$shapiro
levene1 <- arrange(phase2.results1$levene.data.new, nr)$levene
p.value.genes <- intersect(which(shapiro_coeff1 > p.value.shapiro.threshold), 
                           which(levene1 > p.value.levene.threshold))
M.anova.part <- M.anova[p.value.genes,] 
M.anova.used <- cbind(matrix(p.adjust(as.vector(M.anova.part[,1:3]), method= "holm"), ncol=3),
                      M.anova.part[,4]) 
M.anova.used <- as.data.frame(M.anova.used)
colnames(M.anova.used) <- c("cancer.pvalues", "sex.pvalues", "interaction.pvalues", "nr")
```

We would like to show names of the genes that have interactions with sex.
```{r}
interaction.genes <- which(M.anova.used[,3]<= interaction.p.value)
colnames(dataset)[interaction.genes+1]
```

We decided to present one special case for the gene called C20orf134. 
```{r, fig.height=4, fig.width=10}
i <- 1891
d <- filtered.dataset(i)
par(mfrow=c(1,1), mar=c(2,2,2,9))
interaction.plot(d$sex, d$x, d$y, col=rainbow(12), lwd=3, main=paste0(colnames(dataset)[1891]), trace.label="Type of cancer")
```

This case is interesting because for the model which doesn't take sex into consideration we obtained that Rectal Cancer is in its own group, whereas for the model for male only the Rectal Cancer is in the group with Colon Cancer. 
```{r}
model <- lm(y~x, data=d)
LSD.test(model, "x")$groups
model.male <- lm(y ~ x , data=subset(d, sex=="MALE"))
LSD.test(model.male, "x")$groups
```

###Final conclusions 
First observation is that only in few cases cancer types had significant interactions with sex. However, if the interaction exists it has to be considered because results for different genders may be different as it was shown in one of our examples. We managed to find small list of genes that may be used to differentiate all types of genes from each other. It's worth mentioning that for all chosen genes the assumptions of ANOVA hold. 