---
title: "PD3"
author: "Marta Pa³uba"
date: "Modele liniowe i mieszane"
output: 
  html_document:
  toc: TRUE
---
  
### 1. Simulate two datasets, 1000 observations each. Create the first dataset in a way to pass all diagnostic plots (so, all assumptions are valid). Create the second dataset in order to fail with at least 3 diagnostic datasets (e.g. with high Cook distances, non homogeneous variance of residuals, so on).
### 2. Show diagnostic plots for both simulated datasets.

Generujê pierwszy zbiór danych:

```{r}
set.seed(18)
x=rexp(1000)
y=x+rnorm(1000,0,1)
dane1=data.frame(x,y)
```
Zobaczmy jak wygl¹da na wykresach:
```{r}
model1<-lm(y~x, data=dane1)
par(mfrow=c(2,2))
plot(model1)

CD=cooks.distance(model1)
hist(CD,breaks=50)
hist(rstandard(model1),breaks=50)
hist(rstudent(model1),breaks=50)
```

Widaæ, ¿e wszystko siê zgadza.
Mo¿na dodatkowo sprawdziæ rozk³ad normalny reszt testem Shapiro-Wilka
```{r}
shapiro.test(rstandard(model1))
```
pvalue jest wysokie, wiêc nie ma podstaw do odrzucenia hipotezy o normalnoœci rozk³adu reszt.

Generujê drugi zbiór danych:
```{r}
set.seed(18)
x2=rexp(1000)
y21=2*x2[1:500]+rnorm(500,0,1)
y22=17*x2[501:1000]+rnorm(500,0,1)
w=c(y21,y22)
dane2=data.frame(x2,w)
```
Zobaczmy jak wygl¹da na wykresach:

```{r}
model2<-lm(w~x2, data=dane2)
par(mfrow=c(2,2))
plot(model2)

CD2=cooks.distance(model2)
hist(CD2,breaks=50)
hist(rstandard(model2),breaks=50)
hist(rstudent(model2),breaks=50)
```

Widaæ, ¿e teraz ¿e model2 nie zgadza siê na wykresach. Rozk³ad reszt nie jest normalny (na qqplocie ogony znacznie odstaj¹, na histogramach widaæ, ¿e ogony s¹ za d³ugie) ani nie ma homogenicznoœci wariancji.

### 3. Choose three tests for model diagnostic

* Breusch-Godfrey test for higher-order serial correlation.
* Breusch-Pagan test against heteroskedasticity
* Ramsey's RESET test for functional form

### 4. For each test create two simulated datasets. One that passes the given test and one that does not
### 5. Present results from diagnostic tests, show p-values for both datasets.
Skorzystam z danych wygenerwoanych w punkcie 1.

* Breusch-Godfrey test
```{r}
library(lmtest)

bgtest(model1)
bgtest(model2)
```

* Breusch-Pagan test
```{r}
bptest(model1)
bptest(model2)
```

* Ramsey's RESET test
```{r}
resettest(model1)
resettest(model2)
```

Wszêdzie model1 (dla dane1) 'przechodzi' test, a model2 (dla dane2) nie.