---
title: "Praca domowa 3"
author: "Agnieszka Sitko"
date: '2015-10-15'
output:
      html_document:
      toc: yes
pdf_document:
      toc: yes
---
```{r, warning = FALSE, message = FALSE}
library(lmtest)
set.seed(3)
```

<h3>PART I</h3>


`model11` will pass all diagnostic tests. 
```{r, warning = FALSE, message = FALSE}
x11 <- sample(1000)
epsilon11 <- rnorm(1000)
y11 <- x11 + epsilon11
model11 <- lm(y11~x11)
plot(x11, y11)
par(mfrow = c(3,2))
plot(model11, which = 1:6)

#Breusch-Pagan test for homoskedasticity
bptest(model11)

#Goldfeld-Quandt test for homoskedasticity
gqtest(model11)

#Harrison-McCabe test for homoskedasticity
hmctest(y11~x11, order.by=~x11, plot=FALSE,  data = as.data.frame(cbind(x11,y11)))

#Durbin-Watson test for autocorrelation
dwtest(y11~x11, order.by=~x11,  data = as.data.frame(cbind(x11,y11)))

#Breusch-Godfrey test for higher-order serial correlation
bgtest(y11~x11, order.by=~x11,  data = as.data.frame(cbind(x11,y11)))

#Rainbow test for linearity
raintest(y11~x11, order.by=~x11, data = as.data.frame(cbind(x11,y11)))

#Ramsey's RESET test for functional form
resettest(model11, type = "regressor")

#Shapiro-Wilk Normality Test
shapiro.test(rstandard(model11))

```


`model12` will fail at least 3 diagnostic tests.
```{r, warning = FALSE, message = FALSE}
x12 <- 1:1000
epsilon12 <- c(runif(500, -0.5, 0.5), runif(500, -2, 2))
y12 <- log(x12) + epsilon12
model12 <- lm (y12 ~ x12)
par(mfrow = c(1,1))
plot(x12,y12)
abline(model12)
```





It is highly unlikely for `model12` to pass linearity tests. What is more, variance of residuals is suspected to be heterogeneous.
```{r, message=FALSE, warning=FALSE}
par(mfrow = c(3,2))
plot(model12, which = 1:6)
```


Normal QQ plot shows that residuals may not be normally distributed.
```{r, message=FALSE, warning=FALSE}
#Breusch-Pagan test for homoskedasticity
bptest(model12)

#Goldfeld-Quandt test for homoskedasticity
gqtest(model12)

#Harrison-McCabe test for homoskedasticity
hmctest(y12~x12, order.by=~x12,plot=FALSE)

#Durbin-Watson test for autocorrelation
dwtest(y12~x12, order.by=~x12)

#Breusch-Godfrey test for higher-order serial correlation
bgtest(y12~x12, order.by=~x12)

#Rainbow test for linearity
raintest(y12~x12, order.by=~x12)

#Ramsey's RESET test for functional form
resettest(model12, type = "regressor")

#Shapiro-Wilk Normality Test
shapiro.test(rstandard(model12))
```


So, in fact `model12` do not pass any of analyzed diagnostic test.

<h3>PART II</h3>

In each case `model21i` will pass diagnostic test and `model22i` will fail.

<b> Goldfeld - Quandt test for homoskedasticity </b>
```{r, warning = FALSE, message = FALSE}
x21a <- 1:1000
epsilon21a <- rnorm(1000)
y21a <- 1:1000 + epsilon21a
model21a <- lm(x21a~y21a)
plot(x21a,y21a)
abline(model21a)
plot(model21a, which = 1)
gqtest(model21a)


x22a <- 1:1000
epsilon22a <- c(rnorm(100, 0, 10), rnorm(500, 0, 15), rnorm(400, 0, 543))
y22a <- 1:1000 + epsilon22a
model22a <- lm(x22a~y22a)
plot(x22a,y22a)
abline(model22a)
plot(model22a, which = 1)
gqtest(model22a)

```

<b>Durbin-Watson for autocorrelation</b>
```{r, warning = FALSE, message = FALSE}
x21b <- 1:10
y21b <- 1:10
par(mfrow = c(1,1))
plot(x21b,y21b)
model21b <- lm(y21b~x21b)
dwtest(y21b ~ x21b, order.by =~x21b)


x22b <- 1:1000 
y22b <- 1:1000
par(mfrow = c(1,1))
plot(x22b,y22b)
model22b <- lm(y22b ~ x22b)
dwtest(y22b ~ x22b, order.by =~x22b)

```
Although both models in theory are perfectly fitted (which means that fitted values should equal observed values), only first passes tests for autocorrelation. The above example shows that given test of autocorrelation may yield to different results depending on the size of our sample. It is due to the roundings errors which occur while calculating residuals.

<b>RESET test </b>
```{r, warning = FALSE, message = FALSE}
x21c <- sample(1000)
y21c <- x21c
par(mfrow = c(1,1))
plot(x21c,y21c)
model21c <- lm(y21c~x21c)
resettest(model21c)

x22c <- sample(1000)
y22c <- log(x22c)
par(mfrow = c(1,1))
plot(x22c,y22c)
model22c <- lm(y22c ~ x22c)
abline(model22c)
resettest(model22c)

```
