---
title: "Homework 3"
author: "Ma³gorzata Pujszo"
date: "Wednesday, October 21, 2015"
output: html_document
---
1.and 2.

Good dataset and diagnostic plot for it.

```{r warning=FALSE, message=FALSE}
set.seed(9)

x <- runif(1000)
y <- 5*x + 3 + rnorm(1000)
good <- data.frame(y = y,
                  x = x)
model.good <- lm(y~x, data=good)
par(mfrow=c(3, 2))
plot(model.good, which=1:6)
```

Bad dataset and diagnostic plot for it.

```{r warning=FALSE, message=FALSE}
x <- runif(1000, 0, 10)
x[1000]<-20
y[1:999] <- 5000*x[1:999] + (6*x[1:999]-5)^2 + rnorm(999)*exp(x[1:999])
y[1000]<- 152000

bad <- data.frame(y = y,
                  x = x)
model.bad <- lm(y~x, data=bad)
par(mfrow=c(3, 2))
plot(model.bad, which=1:6)
```

As we can see our dataset doesn't pass Scale-Location, Normal Q-Q, and Cook's distance since there exists observation for which Cook's distance > 1.

3. 
I choose tests:
* Shapiro-Wilk Normality Test
* Ramsey's RESET test for functional form
* Harrison-McCabe test for heteroskedasticity

4. 

Datasets for Shapiro-Wilk Normality Test

```{r warning=FALSE, message=FALSE}
x <- runif(1000)
y <- 5*x + 3 + rnorm(1000)
good <- data.frame(y = y,
                  x = x)
model.good <- lm(y~x, data=good)

y <- 5*x + 3 + rgamma(1000, shape=3)
bad <- data.frame(y = y,
                   x = x)
model.bad <- lm(y~x, data=bad)

shapiro.test(rstandard(model.good))
shapiro.test(rstandard(model.bad))
```

As we can see the first example passes the test whereas the second not.

Datasets for Ramsey's RESET test for functional form

```{r warning=FALSE, message=FALSE}
library(lmtest)
x <- runif(1000)
y <- 5*x + 3 + rnorm(1000)
good <- data.frame(y = y,
                  x = x)

y <- exp(x) + log(x) + x + rnorm(1000)
bad <- data.frame(y = y,
                   x = x)

resettest(y~x, data=good, type="regressor")
resettest(y~x, data=bad, type="regressor")
```

Datasets Harrison-McCabe test for heteroskedasticity

```{r warning=FALSE, message=FALSE}
library(lmtest)
x <- runif(1000)
y <- 5*x + 3 + rnorm(1000)
good <- data.frame(y = y,
                  x = x)

y[1:250] <- 10*x[1:250] + rnorm(250)*x[1:250]
y[251:500] <- 10*x[251:500] + rnorm(250, 0, 14)*x[251:500]
y[501:750] <- 10*x[501:750] + rnorm(250, 0, 4)*x[501:750]
y[751:1000] <- 10*x[751:1000] + rnorm(250, 0, 22)*x[751:1000]
bad <- data.frame(y = y,
                   x = x)

hmctest(y~x, data=good, order.by=~x, plot=TRUE)
hmctest(y~x, data=bad, order.by=~x, plot=TRUE)
```

