---
title: "hm3"
author: "Rafał Cylwa"
date: "21 października 2015"
output: html_document
---
```{r}
library(lmtest)
```

1. Simulate two datasets, 1000 observations each. Create the first dataset in a way to pass all diagnostic plots (so, all assumptions are valid). Create the second dataset in order to fail with at least 3 diagnostic datasets (e.g. with high Cook distances, non homogeneous variance of residuals, so on).


Firstly, we generate data passing all diagnostic tests, that is:


```{r}
set.seed(21)
a<-runif(1000, min=-1, max = 1)
b <- a + rnorm(1000, mean = 0, sd = 0.1)
model1 <- lm(b~a)
plot(b~a)
```

Now we execute some tests.

```{r}
bptest(model1)
gqtest(model1)
hmctest(model1)
dwtest(model1, order.by = ~a)
bgtest(model1, order.by = ~a)
raintest(model1)
resettest(model1)
shapiro.test(rstandard(model1))
```
As we expected, all test are passed.

Let's generate second data set.
```{r}
set.seed(21)
c<-seq(from =0, to = 5, length.out = 1000)
d <- exp(c)+ c(runif(500, min = -0.2, max = 0.2), rnorm(500, mean = 0, sd=2))
model2 <- lm(d~c)
plot(d~c)
```

And test it:
```{r}
bptest(model2)
gqtest(model2)
hmctest(model2)
dwtest(model2, order.by = ~c)
bgtest(model2, order.by = ~c)
raintest(model2)
resettest(model2)
shapiro.test(rstandard(model2))
```
This time, none of the tests is passed.

2.Show diagnostic plots for both simulated datasets.

Diagnostic plots for first data set.

```{r}
par(mfrow = c(2,3))
plot(model1, which = 1:6)
```

And for the second one.

```{r}
par(mfrow = c(2,3))
plot(model2, which = 1:6)
```


3-5. In 1. we simulated a first data set that passes all tests and second one that fails all of them. So all required tasks in 3-5. are already done.
