---
title: "Praca domowa 9"
author: "Agnieszka Sitko"
output:
  html_document:
    toc: yes
---
In both scenarios presented below we create matrix Z illustrating a variable which has 100 levels. We consider this variable to have random effect on the dependent variable.

##Scenario I

All random effects are equal to zero.

```{r, warning = FALSE, message = FALSE}
library(lme4)
library(ggplot2)
set.seed(35)

N <- 1000
M <- 1000
noOfGroups <- 100
standardDeviation1 <- 0

Z <- factor(sample(1:noOfGroups, replace = TRUE, size = N))

sigmasScenario1 <- replicate(M,{
      epsilon <- rnorm(N)
      u <- rnorm(noOfGroups, sd = standardDeviation1)
      y <- epsilon + model.matrix( ~ Z -1) %*% u
      model <- lmer(y ~ (1|Z))
      sqrt(VarCorr(model)[[1]][[1]])
})

qplot(sigmasScenario1, geom = "histogram") + ggtitle("Sigma_1 square distribution ")
summary(sigmasScenario1)
```

##Scenario II

$\sigma_1^2 = 1$

```{r, warning = FALSE, message = FALSE}
N <- 1000
M <- 1000
noOfGroups <- 100
standardDeviation2 <- 1

Z <- factor(sample(1:noOfGroups, replace = TRUE, size = N))

sigmasScenario2 <- replicate(M,{
      epsilon <- rnorm(N)
      u <- rnorm(noOfGroups, sd = standardDeviation2)
      y <- epsilon + model.matrix( ~ Z -1) %*% u
      model <- lmer(y ~ (1|Z))
      sqrt(VarCorr(model)[[1]][[1]]) 
})

qplot(sigmasScenario2, geom = "histogram") + ggtitle("Sigma_1 square distribution ")
summary(sigmasScenario2)
```

It looks like in this scenario estimator of $\sigma_1^2$ is continuously distributed on a finite interval. 

##Conclusions:

- in both cases sample means of $\sigma_1^2$'s estimators are close to the real values of the parameters.
- in the first scenario estimated $\sigma_1^2$ has an atom at zero (real value), while in the second scenario its distribution is continuous (thus, does not have any atoms).
- to conclude, the distribution of the estimator of $\sigma_1^2$ is unequivocal and varies with respect to the value of the real parameter.