---
title: "Homework 9"
author: "Ma³gorzata Pujszo"
date: "Monday, December 07, 2015"
output: html_document
---

###Dataset simulation 

Firstly, I created a function which will be used in 2 simulated scenarios. u has normal distribution with mean=0 and standard deviation=sigma1, errors have normal distribution with mean=0 and standard deviation=sigma0. Function sim simulates a dataset, creates a model and returns estimator for sigma1. z is the variable which is modeled as random effect. 

```{r, message=FALSE, warning=FALSE}
library(lme4)
library(ggplot2)
set.seed(77)
sim <- function (z, sigma0, sigma1) {
  epsilon <- rnorm(length(z), 0, sigma0)
  u <- rnorm(length(z), 0, sigma1)
  y <- u[z] + epsilon
  model <- lmer(y~(1|z))
  as.data.frame(VarCorr(model))[1, 4]
}
```

For both scenarios I used the same z, sigma0=2.

###First scenario

Here, sigma1=0. Since I'm looking for distribution of estimator of sigma1^2, I performed sim function n=1000 times. I plotted histogram and compute some basic statistics for the obtained distribution.


```{r}
n <- 1000
z <- rep(1:10, times=1:10)
sigma0 <- 2
sigma1 <- 0

sigma2.1 <- replicate(n, sim(z, sigma0, sigma1))
qplot(sigma2.1^2, stat="ecdf", geom="step") + ggtitle("Distribution function of sigma1^2")
summary(sigma2.1^2)
f1 <- ecdf(sigma2.1^2)
f1(0)
```

As we can see this distribution is skewed with the maximum value around 4 but it is mostly concentrated near 0. Moreover, probability of 0 is around 0.5.

###Second scenario

In this case sigma1=1.

```{r}
sigma1 <- 1
sigma2.2 <- replicate(n, sim(z, sigma0, sigma1))
qplot(sigma2.2^2, stat="ecdf", geom="step") + ggtitle("Distribution function of sigma1^2")
summary(sigma2.2^2)
f2 <- ecdf(sigma2.2^2)
f2(0)
```

It is also a skewed distribution but here it is more spread and probability of 0 is less than 0.1.

###Comparison of 2 scenarios

Let's compare those distributions.

```{r}
df <- data.frame(x = c(sigma2.1^2, sigma2.2^2), scenario = gl(2, n))

ggplot(df, aes(x, colour = scenario)) + stat_ecdf() + ggtitle("Comparison of distribution functions in 2 scenarios")
```


